---
title: "Gene-Environment Correlations and Assortative Mating"
subtitle: "ESSGN 2026 Workshop"
author: "Espen Eilertsen & Hans Fredrik Sunde"
date: 2026-03-25
format:
  revealjs:
    embed-resources: true
    theme: simple
    slide-number: true
    center: false
    chalkboard: false
    smaller: true
    scrollable: true
    transition: fade
    fig-align: center
    width: 1280
    height: 720
    margin: 0.04
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---


## What does a GWAS estimate?
The infensimal model says that the phenotype is a sum of many small genetic effects plus an environment:
$$
y_i = \sum_{j=1}^{M} \beta_j x_{ij} + e_i
$$

In a GWAS, each SNP $j$ is tested separately in a linear model:

$$
y_i = \alpha_j + \tilde{\beta}_j x_{ij} + \dots + \varepsilon_{ij}
$$


$\tilde{\beta}_j$ is the **marginal coefficient** measuring the ***association*** between $x_{ij}$ and $y_i$: 

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\color{red}{\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}}
& + &
\color{blue}{\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}} \\
& &
\scriptstyle {\text{Causal Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

::: {.callout-note appearance="minimal"}
Here we ignore sampling error, imperfect genetic overlap between GWAS and target phenotypes, and heterogeneity across GWAS samples
:::

## Lecture Outline

The unifying theme today is that different sources of confounding ultimately comes down to the same thing: **marginal associations can be confounded by correlation structure**.

1. **Part 1: Local linkage disequilibrium (LD)**

   Why nearby correlated SNPs both distort and inform marginal effect estimates
2. **Part 2: Assortative mating and long-range LD**

   How mating behavioirs affect correlations between distant loci
3. **Part 3: Gene-environment correlations**

   How indirect effects, stratification, and correlated environments enter SNP estimates and variance decompositions
4. **Part 4 (*Maybe*): Family-based methods**

   How family designs remove much of the family-level confounding around marginal SNP estimates
5. **Part 5: Practical**

   Practical 1: GREML and maternal indirect genetic effects with M-GCTA
   Practical 2: SNP heritability and a genetic correlation with LD score regression

# Part 1: Local Linkage Disequilibrium {background-color="#2c3e50" style="color: white;"}

## Nearby SNPs correlate

Recombination shuffles **chromosome segments**, not isolated SNPs. Loci that are close together are less likely to be separated by recombination and therefore tend to be inherited together. 

<table style="border-collapse: collapse; font-family: Lato;">
  <tr><td colspan="11" style="height:10px;"></td></tr>
   <tr>
    <td style="border: none; text-align: right;"><b>$t_{grandmother}$</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">1</td>
  </tr>

  <tr>
    <td style="border: none; text-align: right;"><b>$t_{grandfather}$</b></td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
  </tr>

  <tr><td colspan="11" style="height:10px;"></td></tr>

  <tr>
    <td style="border: none; text-align: right;"><b>$t_{parent}$</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
  </tr>
   <tr><td colspan="11" style="height:10px;"></td></tr>

</table>


This makes nearby SNPs correlated in the population: $\text{Cov}(x_j, x_k) \neq 0$ for nearby $j,k$

This is called (local) **linkage disequilibrium (LD)**.

::: {.callout-note}
If two nearby SNPs are often inherited together, they are statistically *dependent*: once you know the allele at one SNP, you can make a good guess about the allele at the other.
:::


## A two-SNP toy example


Assume the true model is:

$$
y_i = \beta_1 x_{i1} + e_i
$$

where $x_1$ is causal and is in LD with $x_2$, a non-causal SNP:

$$
\beta_1 \neq 0, \qquad \beta_2 = 0, \qquad \text{Cov}(x_1,x_2) \neq 0
$$

A GWAS tests each SNP separately. For $x_2$, the marginal coefficient is:

$$
\tilde{\beta}_2
=
\color{grey}{\beta_2}
\color{grey}{+}
\beta_1 \frac{\text{Cov}(x_2,x_1)}{\text{Var}(x_2)}
$$

In other words, $x_2$ will be associated with the phenotype even though it has **no direct causal effect**.


## Generalizing to many SNPs

Recall that the marginal coefficient at SNP $j$ is (ignoring the rGE term for now):

$$
\tilde{\beta}_j = \beta_j
+ \sum_{k \ne j} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
$$

If we standardize the SNPs so that $\text{Var}(x_j)=1$, then $\text{Cov}(x_j,x_k)=r_{jk}$ and this becomes

$$
\tilde{\beta}_j = \sum_{k=1}^M r_{jk}\beta_k
$$

So the marginal coefficient at SNP $j$ is a weighted sum of the direct effects at correlated SNPs (including itself). A SNP that is correlated with **many** nearby variants can therefore pick up more tagged signal.



## We can exploit this structure {.huge background-color="#5b0f0f" color="white .center}

<br>

If a trait is heritable, then SNPs correlated with many nearby variants should have inflated test statistics on average, because they tag more causal variation in the region.

## What is an LD score?

For SNP $j$, the **LD score** is the sum of squared correlations with nearby SNPs:

$$
\ell_j = \sum_{k \in \mathcal{N}(j)} r_{jk}^2
$$

where $\mathcal{N}(j)$ is a local window around SNP $j$. 

::: {.callout-tip}
# Interpretation
A high LD score means the SNP is highly correlated with many nearby variants, and is more likely to tag causal variation in the region. 

A low LD score means the SNP is more isolated and less likely to tag nearby causal effects.
:::

## LD score regression

If SNPs with higher LD scores tag more nearby causal variation, then under polygenicity their test statistics should be larger on average. This relationship is proportional to the heritability. We can therefore exploit this structure to estimate SNP heritability from GWAS summary statistics, using a method called **LD score regression (LDSC)**.

$$
E[\chi_j^2 \mid \ell_j] = 1 + \frac{N h^2}{M}\ell_j + c
$$

where $\chi_j^2 = z_j^2$, and $z_j = \hat{\tilde{\beta}}_j/\text{SE}(\hat{\tilde{\beta}}_j)$ for SNP $j$.

::: {.callout-tip}
## High-level interpretation
- Regress GWAS test statistics on LD scores
- The **slope** tracks polygenic signal, but it must be rescaled to get heritability:
  $h^2 = \text{slope} \times M / N$
- The **intercept**, $c$, tracks confounding or other inflation

Other considerations include  appropriate regression weights to handle LD-dependent redundancy, plus a well-matched LD reference panel, careful GWAS QC, and checking for sample overlap or ancestry mismatch.
:::

## Genetic correlation from cross-trait LDSC {.smaller}

For two traits, the same idea extends to the product of the two marginal $z$-statistics:

$$
E[z_{1j} z_{2j} \mid \ell_j]
=
0 + 
r_g \,\ell_j
\left(
\frac{\sqrt{N_1 N_2}}{M} \, h_{1} h_{2}
\right)
+ c_{12}
$$

where $r_g$ is the **genetic correlation**, summarizing how strongly the additive genetic effects overlap across the two traits.

::: {.callout-tip}
## High-level interpretation
- Regress cross-trait $z$-score products on LD scores
- The slope tracks **shared genetic signal** across the two traits, but it first has to be rescaled to get the genetic covariance:
  $\mathrm{Cov}(g_1,g_2) = \text{slope} \times M / \sqrt{N_1N_2}$
- Then standardize that covariance to get the genetic correlation:
  $r_g = \frac{\mathrm{Cov}(g_1,g_2)}{h_{1} h_{2}}$
:::

## Summary: local LD both confounds and helps

| Step | Main lesson |
|---|---|
| Nearby SNPs correlate | Recombination makes nearby loci travel together |
| Marginal coefficients absorb tagged effects | So $\tilde{\beta}_j$ need not equal $\beta_j$ |
| LD scores quantify local tagging | They summarize how much nearby variation a SNP captures |
| LD score regression exploits that structure | The same LD that complicates interpretation can be modeled |

::: {.callout-note}
## Reminder:
$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\color{red}{\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}}
& + &
\color{blue}{\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}} \\
& &
\scriptstyle {\text{Causal Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$
:::

# Part 2: Assortative Mating and Long-Range LD {background-color="#2c3e50" style="color: white;"}

## What is assortative mating?

Partner similarity, $\text{Cov}(y_m, y_p) > 0$, can arise for different reasons. 

**Assortative mating (AM)** means that the **matching process itself** induces covariance between partners, rather than that covariance reflecting a shared cause. 

::: {.callout-important}
## Observed partner covariance is a **symptom**, not yet an explanation.

- **Assortative mating**: matching induces partner covariance
- **Convergence**: partners become more similar after pairing
- **Stratification**: partners are similar because both are shaped by the same background factors

Here, we assume that partner covariance is due to assortative mating, and we focus on the genetic consequences of that process.
:::

::: {.callout-note}
## Negative assortative mating?
Technically, assortative mating can be negative if partners are more dissimilar than expected by chance. However, we focus on positive assortative mating here because it is much more common in human populations. The consequences of negative assortative mating are generally the same, but with opposite signs.
:::
## Consequences of assortative mating

::: {.callout-tip}
## Useful intuition
Assortative mating induces correlations among causes by matching on the effects. 
:::


**Ordinary family resemblance**: <br> Covariance starts in the **causes** and is diluted as it shows up in the phenotype 

::: {.callout-note appearance="minimal"}
Correlation flows from $g \to y$.
:::

**Assortative mating**: <br>
Covariance starts on the **phenotype**. If partners are phenotypically similar, then part of that covariance is carried back into the genetic values.

::: {.callout-note appearance="minimal"}
Correlation flows from $y \to g$.
:::


Assortative mating therefore induces genetic similarity between partners.



## Predicted genetic value given phenotype
To quantify the induced genetic similarity, we start at the **phenotype** and ask: if we observe someone's phenotype $y$, what genetic value $g$ do we expect?

Assume centered variables and the simple additive model:

$$
y_i = g_i + e_i;
\quad
\small \text{Cov}(g_i,e_i)=0;
\quad
E[g_i] = E[e_i] = 0
\quad
\Rightarrow \quad
\text{Cov}(g_i,y_i)
= \text{Var}(g_i)
= \sigma_g^2
$$

Here, the regression slope of $g$ on $y$ is

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2}{\sigma_y^2}
= h^2
$$

Meaning the corresponding expected genetic value given phenotype is

$$
E[g_i \mid y_i] = h^2 y_i
$$


::: {.callout-warning}
## $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope even if $\text{Cov}(g,e) \neq 0$, but it no longer equals $h^2$.
<details>
<summary>Show the algebra</summary>
If $\text{Cov}(g,e)\neq 0$, the regression coefficient $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope, but that coefficient is no longer exactly narrow-sense heritability:

If $y=g+e$, then

$$
\text{Cov}(g,y)
= \text{Cov}(g, g+e)
= \text{Var}(g) + \text{Cov}(g,e)
$$

and

$$
\text{Var}(y)
= \text{Var}(g+e)
= \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

Hence

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2 + \sigma_{ge}}{\sigma_g^2 + \sigma_e^2 + 2\sigma_{ge}}
= h^2 + \frac{\sigma_{ge}}{\sigma_y^2}
$$

where $\sigma_{ge} = \text{Cov}(g,e)$ and $h^2 = \sigma_g^2/\sigma_y^2$.

So the regression coefficient equals $h^2$ only when $\text{Cov}(g,e)=0$.
If $\text{Cov}(g,e) > 0$, it is larger than $h^2$; if $\text{Cov}(g,e) < 0$, it is smaller.

</details>
:::

## Quantifying genetic similarity 

Applying $E[g \mid y] = h^2 y$ to both mates, the induced covariance component is:

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}(h^2 y_m, h^2 y_p)
=
h^4 \text{Cov}(y_m,y_p)
$$

So the induced genotypic correlation between mates is

$$
r_g^{\mathrm{mates}}
=
\frac{(\sigma_g^2/\sigma_y^2)^2 \text{Cov}(y_m,y_p)}{\sigma_g^2}
=
\frac{\sigma_g^2}{\sigma_y^2}
\frac{\text{Cov}(y_m,y_p)}{\sigma_y^2}
= h^2 r_y^{\mathrm{mates}}
$$

::: {.callout-note}
The $h^4$ appears on the covariance scale because the phenotype-to-genotype regression slope ($\sigma_g^2/\sigma_y^2 = h^2$) is applied **once for each mate**. When we convert covariance to correlation, one power of $\sigma_g^2$ is divided back out, leaving the simpler result $r_g^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$. 
:::


## Direct versus indirect assortative mating {.smaller}

The preceding (and following) derivations assume that partners match directly on the **focal phenotype** $y$. 

In reality, they may match on a related signal or bundle of traits $s$ that correlates with $y$. We call $s$ the **sorting factor**.

::: {style="font-size: 0.8em;"}
| Case | What partners match on | Induced genotypic correlation between partners |
|--|----|---|
| **Direct assortment** | The focal phenotype itself: $s = y$ | $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,y)^2 r_y^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$ |
| **Indirect assortment** | A related signal or bundle of traits: $s \neq y$, but $\mathrm{Corr}(s,y) \neq 0$ |  $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,s)^2 r_y^{\mathrm{mates}}$ |
:::


::: {.callout-note}
## Key point
The genotypic correlation between partners depends on **how strongly the sorting factor $s$ correlates with the genetic causes of the focal phenotype**. Indirect assortative mating can imply **more or less** genetic resemblance between partners than you would otherwise expect.

In this lecture, we assume direct assortment where $s=y$, but note that assortment is indirect for many social traits (e.g., educational attainment).
:::

## Genetic similarity spreads across many SNP pairs

The genetic value is a weighted sum of SNPs: $g_i = \sum_{j=1}^{M} \beta_j x_{ij}$. Then partner genetic covariance therefore expands as

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}\!\left(\underbrace{\sum_{j=1}^{M} \beta_j x_{mj}}_{g_m}\ ,\  \underbrace{\sum_{k=1}^{M} \beta_k x_{pk}}_{g_p}\right)
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
 \beta_j\beta_k \,\text{Cov}(x_{mj}, x_{pk})
$$

Phenotype matching therefore induces many small SNP-level correlations across partners. Importantly, all these correlations are **positive** for trait-associated SNPs, because the matching process aligns the genetic values of the partners.

::: {.callout-note}
To calculate $\text{Cov}(x_{mj}, x_{pk})$, we can use the same regression logic as before where we regress the two SNPs on their phenotype and combine that with the phenotypic covariance: $\frac{\text{Cov}(x_{mj}, y_m)}{\text{Var}(y_m)} \text{Cov}(y_m, y_p) \frac{\text{Cov}(y_p, x_{pk})}{\text{Var}(y_p)}$.
:::


## Why is this a problem? {.huge background-color="#5b0f0f" color="white" .center}
::: {.callout-important  appearance="simple"}
Once partner genomes become correlated, transmission and recombination **embeds the correlation structure** in the next following generation. 

Assortative mating across generations induces a signed correlation structure between trait-associated alleles (i.e., linkage disequilibrium), and that correlation structure can increase marginal SNP effects and family resemblance.

Now we will unpack how that happens.
:::

## What happens after *one* generation of AM 

First, we define the genetic values separately for the maternal and paternal haplotypes:

$$
g_c^{(m)} = \sum_{j=1}^{M} \beta_j t_{mj}, 
\qquad
g_c^{(p)} = \sum_{j=1}^{M} \beta_j t_{pj},
\qquad
g_c = g_c^{(m)} + g_c^{(p)}
$$

where $t_{mj}$ and $t_{pj}$ are the maternally and paternally transmitted alleles at SNP $j$, respectively. 

After only one generation of assortment, the transmitted alleles are still independent **within** haplotypes (assuming no relevant local LD):

$$
\operatorname{Cov}(t_{m j}, t_{m k}) = 0,
\qquad
\operatorname{Cov}(t_{p j}, t_{p k}) = 0
\qquad \text{for } j \neq k
$$

However, the transmitted alleles are correlated **across** haplotypes, because the parental genomes are correlated: 

$$
\text{Cov}(g_c^{(m)}, g_c^{(p)}) > 0
\qquad
\Rightarrow
\qquad
\text{Cov}(t_{mj}, t_{pk}) > 0 \small \text{ for trait-associated } j,k
$$


## Genetic variance after one generation of AM
Because the transmitted alleles are still independent within haplotypes, the variance of each haplotype contribution is just a sum of single-locus terms:
$$
\text{Var}(g_c^{(m)}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(t_{mj}),
\qquad
\text{Var}(g_c^{(p)}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(t_{pj})
$$

However, because the transmitted alleles from either parent are correlated, the total variance of the genetic value includes a positive covariance term:

$$
\text{Var}(g_c)
=
\text{Var}(g_c^{(m)})
+ \text{Var}(g_c^{(p)})
+ 2\text{Cov}(g_c^{(m)}, g_c^{(p)})
$$


::: {.callout-note}
## Terminology: trans vs cis covariance
Covariance **across** the maternal and paternal haplotypes is called **trans** covariance.

Covariance **within** the maternal or paternal haplotype is called **cis** covariance.
:::

## What happens after *several* generations of AM

Repeated assortment plus recombination gradually converts the **trans** covariance into **cis** covariance. 
$$
\text{Cov}(t_{mj}, t_{mk}) \neq 0
\qquad
\text{for distant loci } j,k
$$

That means the haplotype variance is no longer just a sum of single-locus terms. Instead, the variance now includes off-diagonal covariance terms between trait-associated loci. For example,

$$
\text{Var}(g_c^{(m)})
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
\text{Cov}(\beta_j t_{mj}, \beta_k t_{mk})
$$

<br>

::: {style="font-size: 0.8em;"}
| When | Where | Interpretation |
|--|---|-----|
| 1 generation | 	Across haplotypes (trans)  | Maternal and paternal contributions become aligned |
| ≥2 generations | Within haplotypes (cis) | Aligned alleles begin to co-occur within haplotypes |
:::

## Why marginal SNP effects now tag distant variants {.smaller}

Again set the rGE term aside and focus on the LD component:

$$
\tilde{\beta}_j
= \beta_j
+ \sum_{k \ne j} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
$$

Under local LD, the sum is dominated by nearby loci. Under assortative mating, some of those covariance terms can involve **distant** causal variants.

::: {.callout-note}
So a marginal hit at SNP $j$ can partly reflect genome-wide covariance structure, not only local biology at SNP $j$.
:::

## Why $\text{Var}(g)$ increases for a few generations {.smaller}

Using the haplotype decomposition,

$$
\text{Var}(g_c)
=
\text{Var}(g_c^{(m)})
+ \text{Var}(g_c^{(p)})
+ 2\text{Cov}(g_c^{(m)}, g_c^{(p)})
$$

After one generation, the increase comes mainly from the positive **trans** term $2\text{Cov}(g_c^{(m)},g_c^{(p)})$. After later generations, each haplotype contribution also contains positive **cis** terms:

$$
\text{Var}(g_c^{(m)})
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
\text{Cov}(\beta_j t_{mj}, \beta_k t_{mk})
$$

and similarly for $g_c^{(p)}$.

| Generation | Main covariance pattern | Expected $\text{Var}(g)$ |
|---|---|---|
| Random mating baseline | Little long-range covariance | Baseline |
| After 1 generation | Positive trans covariance | Higher |
| After several generations | Trans + cis covariance | Higher still |
| Equilibrium | Assortment balanced by recombination | Plateau |

::: {.callout-note}
This trajectory is schematic: positive assortment tends to increase genetic variance for a few generations and then approach a new equilibrium, not increase forever.
:::

## Family resemblance increases under assortative mating

Once parental genomes are more aligned, resemblance among relatives increases for the trait under assortment.

If mates are positively correlated in genetic value, then:

$$
\text{Cov}(g_m, g_p) > 0
$$

and therefore:

$$
\text{Var}\!\left(\frac{g_m + g_p}{2}\right)
= \frac{1}{4}\text{Var}(g_m)
+ \frac{1}{4}\text{Var}(g_p)
+ \frac{1}{2}\text{Cov}(g_m, g_p)
$$

::: {.callout-note}
The same induced covariance structure now appears in two places:
- in GWAS, through long-range tagging
- in families, through stronger resemblance among relatives
:::

## Summary: assortative mating changes GWAS, LD, and family resemblance

| Step | Main lesson |
|---|---|
| Mates match on phenotype | Start with $\text{Cov}(y_m,y_p) > 0$ |
| First generation | Mostly trans covariance appears |
| Later generations | Cis covariance and long-range LD build up |
| Population consequence | $\text{Var}(g)$ rises for a while, then approaches equilibrium |
| Family consequence | Relatives resemble each other more |

::: {.callout-note}
Again, marginal associations reflect **induced correlation structure**, not only direct causal effects.
:::

# Part 3: Gene-Environment Correlations {background-color="#2c3e50" style="color: white;"}

## Same regression, new covariance sources

Now set the LD term aside and focus on gene-environment correlation:

$$
\tilde{\beta}_j = \beta_j + \frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}
$$

Part 3 is about a new question:

$$
\text{What can generate } \text{Cov}(x_j, e)?
$$

::: {.callout-note}
In Part 1 and Part 2, the extra terms came mainly from **correlated genotypes**. Here the focus is **correlated environments**.
:::

## Where can $\text{Cov}(x_j,e)$ come from?

| Source | Why genotype and environment become correlated |
|---|---|
| Passive rGE | Parents transmit both genes and family environments |
| Indirect genetic effects | Related people's genotypes shape the focal person's environment |
| Stratification / population structure | Allele frequencies and environments differ across groups |
| Residual confounding | Selection, ascertainment, measurement, or omitted structure |

::: {.callout-note}
Different mechanisms can produce the **same statistical symptom** in a marginal GWAS coefficient: $\text{Cov}(x_j,e) \neq 0$.
:::

## Indirect genetic effects

A focal person's phenotype can be affected not only by their own genotype, but also by the genotypes of parents, siblings, peers, or partners through the environments those people create.

$$
y_i = g_i^{\text{direct}} + g_i^{\text{indirect}} + e_i
\qquad
g_i^{\text{direct}} = \sum_{j=1}^{M}\beta_j x_{ij}
$$

:::: {.columns}
::: {.column width="50%"}
```{dot}
//| fig-width: 5
//| fig-height: 3.1
//| echo: false
digraph ige {
  rankdir=LR;
  graph [margin=0.08, nodesep=0.45, ranksep=0.7];
  node [shape=circle, style=filled, fixedsize=true, width=0.85, height=0.85,
        fontsize=17, fontcolor=white, fontname="Helvetica-Bold", penwidth=2];
  edge [penwidth=2, fontsize=15, fontname="Helvetica-Bold"];

  xr [label="x_rel", fillcolor="#e74c3c", color="#c0392b"];
  env [label="env", fillcolor="#e67e22", color="#ca6f1e"];
  xi [label="x_i", fillcolor="#3498db", color="#2980b9"];
  yi [label="y_i", fillcolor="#7f8c8d", color="#707b7c"];

  xr -> env [label=" indirect"];
  env -> yi [label=" environment"];
  xi -> yi [label=" direct"];
}
```
:::
::: {.column width="50%"}

- Your own genotype can affect your phenotype directly
- Other people's genotypes can affect your phenotype indirectly by shaping your environment
- That is why indirect genetic effects are not the same thing as stratification

:::
::::

## Stratification and population structure

Stratification can also induce correlation between genotype and environment, but through group-level differences rather than through other people's genotypes creating the focal environment.

```{dot}
//| fig-width: 5.2
//| fig-height: 3
//| echo: false
digraph strat {
  rankdir=TB;
  graph [margin=0.08, nodesep=0.5, ranksep=0.7];
  node [shape=circle, style=filled, fixedsize=true, width=0.9, height=0.9,
        fontsize=17, fontcolor=white, fontname="Helvetica-Bold", penwidth=2];
  edge [penwidth=2, fontsize=15, fontname="Helvetica-Bold"];

  anc [label="group", fillcolor="#8e44ad", color="#7d3c98"];
  xj  [label="x_j", fillcolor="#3498db", color="#2980b9"];
  e   [label="e", fillcolor="#e67e22", color="#ca6f1e"];
  y   [label="y", fillcolor="#7f8c8d", color="#707b7c"];

  anc -> xj;
  anc -> e;
  xj -> y [label=" marginal"];
  e  -> y [label=" env"];
}
```

::: {.callout-important}
Stratification is **confounding**, not nurture. It can create $\text{Cov}(x_j,e)$ even when there is no within-family environmental pathway from genotype to phenotype.
:::

## Back to the marginal SNP coefficient {.smaller}

For this part, the relevant simplification is:

$$
\tilde{\beta}_j
= \beta_j + \frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}
$$

::: {.callout-note}
The orange term can now reflect many different things:
- passive, evocative, or active gene-environment correlation
- indirect genetic effects from parents, siblings, peers, or partners
- stratification and population structure
- other residual confounding
:::

## From SNP confounding to variance decomposition

For the next two slides, let $g$ denote the focal individual's **direct additive genetic value**:

$$
g_i = \sum_{j=1}^{M} \beta_j x_{ij}
$$

Then:

$$
y = g + e
\qquad \Longrightarrow \qquad
\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

::: {.callout-note}
If you split indirect genetic effects out as their own term, you get more covariance terms, not fewer. The interpretational problem does not disappear.
:::

## Why covariance terms complicate interpretation

Keep these quantities conceptually separate:

$$
\text{Var}(g) \qquad \text{Var}(e) \qquad \text{Cov}(g,e) \qquad \text{Var}(y)
$$

The sign of the covariance matters:

| Sign of $\text{Cov}(g,e)$ | Interpretation |
|---|---|
| Positive | Genetically higher-valued individuals tend to experience environments that push in the same direction |
| Negative | Genetically higher-valued individuals tend to experience environments that push in the opposite direction |

::: {.callout-important}
$\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)$ is a **statistical identity**, not a clean causal partition of phenotypic variation.
:::

## Take-home map

| Quantity | What it means | What it does not automatically mean |
|---|---|---|
| Direct effect $\beta_j$ | Causal effect of locus $j$ in the full model | The same thing as the marginal coefficient |
| Marginal coefficient $\tilde{\beta}_j$ | Association between SNP $j$ and phenotype | A pure direct effect |
| Indirect genetic effect | Other people's genotypes shape the focal environment | Population stratification |
| Stratification | Group structure aligns genotypes and environments | Genetic nurture |
| $\text{Cov}(g,e)$ | Statistical alignment of genetic values and environments | A causal share of variance |

::: {.callout-note}
Across local LD, assortative mating, and gene-environment correlation, the same lesson keeps returning: **marginal regression coefficients are shaped by correlation structure**.
:::

# Part 4: Family GWAS {background-color="#2c3e50" style="color: white;"}

## Why families help

Much of this lecture has been about the fact that the **population** coefficient $\tilde{\beta}_j$ can include more than the direct effect $\beta_j$.

If we compare unrelated people, that coefficient can still reflect LD, assortative mating, stratification, genetic nurture, and other family-level environments.

::: {.callout-note}
Family GWAS changes the comparison. Instead of relying only on differences across unrelated families, it uses variation **within families** to get closer to the direct effect.
:::

## A family-based regression

Index individuals by $i$ within family $f$. A simple within-family model is:

$$
y_{if} = \alpha_f + \tilde{\beta}_j^{F} x_{ijf} + \varepsilon_{if}
$$

where $\alpha_f$ is a family intercept shared by relatives in the family design.

With two siblings, this is equivalent to differencing:

$$
y_{1f} - y_{2f}
= \tilde{\beta}_j^{F}\left(x_{1jf} - x_{2jf}\right)
+ \left(\varepsilon_{1f} - \varepsilon_{2f}\right)
$$

::: {.callout-note}
This is exactly where Session 1's **segregation variance** comes back in. The sibling difference $x_{1jf} - x_{2jf}$ is generated by the random segregation deviations in transmission, not by between-family differences in ancestry or parental background.
:::

::: {.callout-note}
Sibling differencing is only **one example** of a family design. The next day's session will go deeper into family-based GWAS more broadly.
:::

## Comparing the two coefficients

| Estimate | What variation identifies it | What it tends to pick up |
|---|---|---|
| Population coefficient $\tilde{\beta}_j$ | Differences across individuals and families | Direct effects plus between-family confounding and tagging structure |
| Family-design coefficient $\tilde{\beta}_j^{F}$ | Segregation-driven differences within families | Often closer to the direct effect $\beta_j$ |

::: {.callout-note}
The point is not that $\tilde{\beta}_j^{F}$ is magically perfect. The point is that it removes much of the **between-family** structure that biases the population coefficient.
:::

## Why this solves many earlier problems

Relatives in the same family share parents, ancestry, and much of their family environment. That means many family-level confounders are held fixed by the comparison.

| Earlier problem | Why family GWAS helps |
|---|---|
| Stratification / population structure | Relatives in the same family share the same family ancestry background |
| Passive rGE / parental genetic nurture | Parental genotypes and much of the rearing environment are common within families |
| Assortative mating background | The parental genomes are the same for the compared relatives |
| Shared family environment | The family intercept absorbs what relatives share |

::: {.callout-note}
So $\tilde{\beta}_j^{F}$ is often **closer to the direct effect** than the population-level marginal coefficient.
:::

## What it does not fully solve

Family GWAS solves **a lot**, but not everything.

| Remaining issue | Why it can remain |
|---|---|
| Local LD | Relatives still inherit correlated nearby variants |
| Indirect genetic effects | One relative's genotype can affect another relative's environment |
| Within-family environmental differences | Family members do not experience identical environments |
| Lower power | Within-family genotype differences are smaller than population-wide variation |

::: {.callout-important}
Family GWAS is best thought of as a design that removes much of the **between-family confounding** around $\tilde{\beta}_j$. It does not magically make every estimate equal the direct causal effect.
:::

## Summary: family GWAS gets closer to direct effects

| Comparison | Main source of identifying variation |
|---|---|
| Population GWAS | Differences across people, families, and environments |
| Family GWAS | Mendelian segregation across relatives within the same family |

::: {.callout-note}
This is why family GWAS is such a useful complement to the rest of the lecture: it directly targets many of the confounding mechanisms we have been discussing, and tomorrow's session will go deeper into the design details.
:::

# Part 5: Practical {background-color="#2c3e50" style="color: white;"}

## Practical 1: GREML and M-GCTA with simulated trios

Run `practical/sim_genomic.R` (or `practical/sim_genomic.jl`) end-to-end:

1. Simulate genotypes for `N` trios at `M` SNPs
2. Simulate phenotypes with direct SNP effects and `h^2 = 0.5`
3. Estimate `h^2` by parent-offspring regression
4. Estimate `h^2` by GREML
5. Simulate an offspring phenotype with a maternal indirect genetic effect
6. Refit the wrong direct-only GREML model
7. Fit the maternal model (M-GCTA) and compare the variance components

::: {.callout-note}
Parts 1-4 match the Session 1 practical. The new Session 2 material is the extension in parts 5-7.

As before, computing the GRM is the slow step. Use that time to discuss what `G` contains and why GREML treats it as the covariance structure for genetic values.
:::

## Part 1: Simulate genotypes {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-sim-genotypes
#| echo: true
#| code-fold: false
#| cache: true
set.seed(080318)
library(gaston)

N = 3000  # Trios
M = 12000 # SNPs
p = 0.5
Xf = matrix(rbinom(N * M, 2, p), N, M)
Xm = matrix(rbinom(N * M, 2, p), N, M)
Xo = matrix(
  rbinom(N * M, 1, as.vector(Xf) / 2) +
  rbinom(N * M, 1, as.vector(Xm) / 2), N, M)
X = rbind(Xf, Xm, Xo)

idf = 1:N
idm = (N + 1):(2 * N)
ido = (2 * N + 1):(3 * N)
```

### Julia
```julia
using LinearAlgebra, Distributions, DataFrames, GLM
using GREMLModels, Random
Random.seed!(080318)

N, M = 3000, 12000
p = 0.5
Xf = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xm = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xo = rand.(Bernoulli.(Xf ./ 2)) .+ rand.(Bernoulli.(Xm ./ 2))
X = vcat(Xf, Xm, Xo)

idf = 1:N
idm = N+1:2N
ido = 2N+1:3N
```

::::

## Part 2: Simulate phenotypes {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-sim-phenotypes
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-sim-genotypes"
p_est = colMeans(X) / 2
X_tilde = scale(X, 2 * p_est, sqrt(2 * p_est * (1 - p_est)))
s2_g = 0.5
s2_e = 0.5
b = rnorm(M, 0, sqrt(s2_g / M))
g = as.vector(X_tilde %*% b)
e = rnorm(3 * N, 0, sqrt(s2_e))
y = g + e

cor(cbind(g[idf], g[idm], g[ido]))
cov(cbind(g[idf], g[idm], g[ido]))
```

### Julia
```julia
p_est = vec(mean(X, dims = 1)) ./ 2
X_tilde = (Float64.(X) .- (2 .* p_est)') ./ sqrt.(2 .* p_est .* (1 .- p_est))'

sigma2_g, sigma2_e = 0.5, 0.5
b = randn(M) * sqrt(sigma2_g / M)
g = X_tilde * b
e = randn(3N) * sqrt(sigma2_e)
y = g + e

cor(hcat(g[idf], g[idm], g[ido]))
cov(hcat(g[idf], g[idm], g[ido]))
```

::::

## Part 3: Parent-offspring regression {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-po-regression
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-sim-phenotypes"
yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
m_po = lm(yo ~ yp)
summary(m_po)
```

### Julia
```julia
yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
df1 = DataFrame(yo = yo, yp = yp)
mpo = lm(@formula(yo ~ yp), df1)
```

::::

## Part 4: GREML direct model {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-greml
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-po-regression"
G = tcrossprod(X_tilde) / M
Goo = G[ido, ido]
m_direct = lmm.aireml(yo, K = Goo, verbose = FALSE)
m_direct$tau
```

### Julia
```julia
G = (X_tilde * X_tilde') / M
Goo = G[ido, ido]
Id = Diagonal(ones(N))
r = [Goo, Id]
mdirect = fit(GREMLModel, @formula(yo ~ 1), df1, r; verbose = false)
```

::::

## Part 5: Simulate a maternal indirect genetic effect {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-maternal-sim
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-greml"
Sg = rbind(
  c(0.2, 0.2),
  c(0.2, 0.5)
)
s2_e = 0.1
B = matrix(rnorm(2 * M), M, 2) %*% chol(Sg / M)
g_m = X_tilde %*% B[, 1]
g_d = X_tilde %*% B[, 2]
e_2 = g_m[idm] + rnorm(N, 0, sqrt(s2_e))
yo_2 = g_d[ido] + e_2
```

### Julia
```julia
Sigma_g = [0.2 0.2; 0.2 0.5]
sigma2_e = 0.1
C = cholesky(Symmetric(Sigma_g / M)).U
B = randn(M, 2) * C
g_m = X_tilde * B[:, 1]
g_d = X_tilde * B[:, 2]
e2 = g_m[idm] + randn(N) * sqrt(sigma2_e)
yo2 = g_d[ido] + e2
```

::::

## Part 6: Fit the wrong direct-only model {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-wrong-direct
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-maternal-sim"
m_direct_2 = lmm.aireml(yo_2, K = Goo, verbose = FALSE)
m_direct_2$tau
```

### Julia
```julia
df2 = DataFrame(y = yo2)
mdirect2 = fit(GREMLModel, @formula(y ~ 1), df2, r; verbose = false)
```

::::

## Part 7: Fit the maternal model (M-GCTA) {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-mgcta
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-wrong-direct"
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
m_maternal = lmm.aireml(yo_2, K = list(Gmm, Dom, Goo), verbose = FALSE)
m_maternal$tau
```

### Julia
```julia
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
r2 = [Gmm, Dom, Goo, Id]
mmaternal = fit(GREMLModel, @formula(y ~ 1), df2, r2; verbose = false)
```

::::

## Interpretation checkpoint

After running the simulation, check that you can answer these questions:

1. Does parent-offspring regression recover something close to the simulated `h^2 = 0.5` in the direct-only setting?
2. In the direct-only GREML model, what do the estimated variance components look like before and after adding the maternal indirect effect?
3. Once you fit the maternal model, do the variance components move closer to the data-generating values?
4. Why does the direct-only model misattribute part of the maternal effect?

## Practical 2: LD score regression with real GWAS data

In this practical, we will use **LD score regression (LDSC)** to estimate:

1. observed-scale SNP heritability for **intelligence**
2. observed-scale SNP heritability for **p_factor**
3. the genetic correlation between the two traits

The summary statistics come from the **CTG summary statistics archive** and the practical uses a local course installer around the newer Python 3 LDSC workflow from **CBIIT/ldsc**. LDSC itself remains the canonical method originally associated with the **bulik/ldsc** project.

- [CTG GWAS summary statistics](https://ctg.thebluebus.nl/software/summary_statistics/)
- [CBIIT/ldsc repository](https://github.com/CBIIT/ldsc)
- [Original bulik/ldsc repository](https://github.com/bulik/ldsc)
- [LDSC heritability and genetic correlation tutorial](https://github.com/bulik/ldsc/wiki/Heritability-and-Genetic-Correlation)

::: {.callout-note}
To keep the exercise focused, the instructor has already prepared helper scripts for the repetitive setup work. You will still run the same main LDSC steps used in practice: `munge_sumstats.py`, single-trait `--h2`, and cross-trait `--rg`, but the course scripts handle the local installation and platform details.
:::

## Setup and inputs {.smaller}

We will work with two pre-harmonized files prepared from the public CTG releases:

- `practical/ldsc/data/course/intelligence_ctg.tsv.gz`
- `practical/ldsc/data/course/p_factor.tsv.gz`

Each file has the same minimal schema:

`SNP`, `A1`, `A2`, `P`, `N`, `Z`

We also need the LDSC reference files:

- `practical/ldsc/data/ref/w_hm3.snplist`
- `practical/ldsc/data/ref/eur_w_ld_chr/`
- `practical/ldsc/data/ref/weights_hm3_no_hla/`

Run the course workflow in this order from the repository root. The practical is intended to work in **Git Bash on Windows**, with the scripts handling the local Python and LDSC setup.

```bash
Rscript practical/ldsc/prepare_ldsc_data.R --force
bash practical/ldsc/install_ldsc.sh
bash practical/ldsc/ldsc_pipeline.sh
```

::: {.callout-note}
These LD scores are based on **1000 Genomes Europeans**, so the exercise assumes European-ancestry GWAS summary statistics. In real work, ancestry mismatch between the GWAS and the reference panel can bias interpretation.
:::

## Part 1: Munge the summary statistics {.smaller}

After running the pipeline above, the first LDSC step is to convert the prepared text files into LDSC's `.sumstats.gz` format. If you want to inspect the script itself, see `practical/ldsc/ldsc_pipeline.sh`.

Look for:

- how many SNPs survive QC and allele matching for `intelligence` and `p_factor`
- whether the munge logs contain warnings
- the two output files `intelligence_ctg.sumstats.gz` and `p_factor.sumstats.gz`

## Part 2: Estimate SNP heritability {.smaller}

That same pipeline run then performs single-trait LDSC for each phenotype.

Look for these lines in each log:

- `Total Observed scale h2`
- `Intercept`
- `Ratio`

The key log files are:

- `practical/ldsc/data/results/intelligence_h2.log`
- `practical/ldsc/data/results/p_factor_h2.log`

::: {.callout-note}
For this short practical we stay on the **observed scale**. We are not doing liability-scale conversion or partitioned heritability here.
:::

## Part 3: Estimate the genetic correlation {.smaller}

Finally, that same pipeline run performs the cross-trait LDSC regression using the two munged `.sumstats.gz` files.

Look for:

- `Genetic Correlation`
- its standard error
- the cross-trait intercept
- the `P:` line

The main log file is:

- `practical/ldsc/data/results/intelligence_p_factor_rg.log`

For these traits, we expect a **negative** genetic correlation.

## Interpretation checkpoint

After the run, check that you can answer these questions:

1. Which trait has the larger observed-scale SNP heritability: `intelligence` or `p_factor`?
2. Are the LDSC intercepts close to 1, or do they suggest substantial inflation beyond polygenicity?
3. Are the LDSC ratios modest, or do they suggest a large share of inflation is due to confounding?
4. Is the estimated genetic correlation between `intelligence` and `p_factor` negative, and how large is it in magnitude?
5. Is the cross-trait intercept close to 0, or could sample overlap / shared confounding be contributing?

::: {.callout-important}
The important habit is not just to report `h^2` and `r_g`, but also to inspect the **intercepts** and **ratios** and ask whether the analysis assumptions look reasonable.
:::





