---
title: "Gene-Environment Correlations and Assortative Mating"
subtitle: "ESSGN 2026 Workshop"
author: "Espen Eilertsen & Hans Fredrik Sunde"
date: 2026-03-25
format:
  revealjs:
    embed-resources: true
    theme: simple
    slide-number: true
    center: false
    chalkboard: false
    smaller: true
    scrollable: true
    transition: fade
    fig-align: center
    width: 1280
    height: 720
    margin: 0.04
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
include-in-header:
  text: |
    <script type="text/x-mathjax-config">
    MathJax.Hub.Config({
      TeX: {
        Macros: {
          mcol: ["{\\color{red}{#1}}", 1],
          pcol: ["{\\color{darkblue}{#1}}", 1]
        }
      }
    });
    </script>
---

## Lecture Outline

::: {.callout-tip}
## The unifying theme
 Different sources of confounding ultimately comes down to the same thing: **unmodelled correlation structure**.

:::

::: {style="font-size: 0.8em;"}
| Part | Focus |
|---|-----|
| **Part 1: Local linkage disequilibrium (LD)** | Why nearby correlated SNPs both distort and inform marginal effect estimates |
| **Part 2: Assortative mating and long-range LD** | How non-random mating affects correlations between distant loci |
| **Part 3: Gene-environment correlations** | How correlations between genetic and environmental factors affect marginal associations and variance decompositions |
| **Part 4: What to do? Use families!** | What family-based methods can and cannot solve |
| **Part 5: Practical** | GREML and maternal indirect genetic effects with M-GCTA |
:::


## Before we begin... {.smaller}

 In session 1, we assumed that loci were independent from each other and from environmental effects. 

$$
\underbrace{\text{Cov}(x_j, e) = 0}_{\text{No rGE}}
\qquad \text{and} \qquad
\underbrace{\text{Cov}(x_j, x_k) = 0}_{\text{No LD}}
\text{ for } j \neq k
$$

This is often not the case.

<br>

**This session: What happens when those assumptions are violated?**

This lecture is mostly **theoretical**. The main takeaway is not a ready-to-use practical method, but a clearer picture of **what happens to target parameters and estimated parameters**.





## Direct effect *vs* Marginal association?
As a reminder, the infinitesimal model says that the phenotype is a sum of many small genetic effects plus a residual (i.e., an environmental effect):
$$
y_i = \sum_{j=1}^{M} \beta_j x_{ij} + e_i
$$

In a GWAS, each SNP $j$ is tested separately in a linear model:

$$
y_i = \alpha_j + \tilde{\beta}_j x_{ij} + \dots + \varepsilon_{ij}
$$


$\tilde{\beta}_j$ is the **marginal coefficient** measuring the ***association*** between $x_{ij}$ and $y_i$: 

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
& + &
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)} \\
& &
\scriptstyle {\text{True Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

::: {.callout-note appearance="minimal"}
Here we ignore sampling error, imperfect genetic overlap between GWAS and target phenotypes, and heterogeneity across GWAS samples
:::

# Part 1: Local Linkage Disequilibrium {background-color="#2c3e50" style="color: white;"}

## Nearby SNPs correlate

Recombination shuffles **chromosome segments**, not isolated SNPs. Loci that are close together are less likely to be separated by recombination and therefore tend to be inherited together. 

<table style="border-collapse: collapse; font-family: Lato;">
  <tr><td colspan="11" style="height:10px;"></td></tr>
   <tr>
    <td style="border: none; text-align: right;"><b>Haplotype 1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">1</td>
  </tr>

  <tr>
    <td style="border: none; text-align: right;"><b>Haplotype 2</b></td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
  </tr>

  <tr><td colspan="11" style="height:10px;"></td></tr>

  <tr>
    <td style="border: none; text-align: right;"><b>Transmitted</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
  </tr>
   <tr><td colspan="11" style="height:10px;"></td></tr>

</table>


This makes nearby SNPs correlated in the population: $\text{Cov}(x_j, x_k) \neq 0$ for nearby $j,k$

This is called (local) **linkage disequilibrium (LD)**.

::: {.callout-note}
If two nearby SNPs are often inherited together, they are statistically *dependent*: once you know the allele at one SNP, you can make a good guess about the allele at the other.
:::


## A two-SNP toy example


Assume the true model is:

$$
y_i = \beta_1 x_{i1} + e_i
$$

where $x_1$ is causal and is in LD with $x_2$, a non-causal SNP:

$$
\beta_1 \neq 0, \qquad \beta_2 = 0, \qquad \text{Cov}(x_1,x_2) \neq 0
$$

A GWAS tests each SNP separately. For $x_2$, the marginal coefficient is:

$$
\tilde{\beta}_2
=
\color{grey}{\beta_2}
\color{grey}{+}
\beta_1 \frac{\text{Cov}(x_2,x_1)}{\text{Var}(x_2)}
$$

In other words, $x_2$ will be associated with the phenotype even though it has **no direct causal effect**.


## Generalizing to many SNPs

Recall that the marginal coefficient at SNP $j$ is (ignoring the rGE term for now):

$$
\tilde{\beta}_j = \beta_j
+ \sum_{k \ne j} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
$$

If we standardize the SNPs so that $\text{Var}(x_j)=1$, then $\text{Cov}(x_j,x_k)=r_{jk}$ and this becomes

$$
\tilde{\beta}_j = \sum_{k=1}^M r_{jk}\beta_k
$$

So the marginal coefficient at SNP $j$ is a weighted sum of the direct effects at correlated SNPs (including itself). A SNP that is correlated with **many** nearby variants can therefore pick up more signal.



## We can exploit this structure {.huge background-color="#5b0f0f" color="white .center}

<br>

If a trait is heritable, then SNPs correlated with many nearby variants should have inflated test statistics on average, because they tag more causal variation in the region.

## What is an LD score?

For SNP $j$, the **LD score** is the sum of squared correlations with nearby SNPs:

$$
\ell_j = \sum_{k \in \mathcal{N}(j)} r_{jk}^2
$$

where $\mathcal{N}(j)$ is a local window around SNP $j$. 

::: {.callout-tip}
# Interpretation
A high LD score means the SNP is highly correlated with many nearby variants, and is more likely to tag causal variation in the region. 

A low LD score means the SNP is more isolated and less likely to tag nearby causal effects.
:::

## LD score regression

If SNPs with higher LD scores tag more nearby causal variation, then under polygenicity their test statistics should be larger on average. This relationship is proportional to the heritability. We can therefore exploit this structure to estimate SNP heritability from GWAS summary statistics, using a method called **LD score regression (LDSC)**.

$$
E[\chi_j^2 \mid \ell_j] = 1 + h^2 \ell_j \frac{N}{M} + c
$$

where $\chi_j^2 = z_j^2$, and $z_j = \hat{\tilde{\beta}}_j/\text{SE}(\hat{\tilde{\beta}}_j)$ for SNP $j$.

::: {.callout-tip}
## High-level interpretation
- Regress GWAS test statistics on LD scores
- The **slope** tracks polygenic signal, but it must be rescaled to get heritability:
  $h^2 = \text{slope} \times M / N$
- The **intercept**, $c$, tracks confounding or other inflation

Other considerations include  appropriate regression weights to handle LD-dependent redundancy, plus a well-matched LD reference panel, careful GWAS QC, and checking for sample overlap or ancestry mismatch.
:::

## Genetic correlation from cross-trait LDSC {.smaller}

For two traits, the same idea extends to the product of the two marginal $z$-statistics:

$$
E[z_{1j} z_{2j} \mid \ell_j]
=
0 + 
r_g \,\ell_j
\left(
\frac{\sqrt{N_1 N_2}}{M} \, h_{1} h_{2}
\right)
+ c_{12}
$$

where $r_g$ is the **genetic correlation**, summarizing how strongly the two traits share genetic signal.

::: {.callout-tip}
## High-level interpretation
- Regress cross-trait $z$-score products on LD scores
- The slope tracks **shared genetic signal** across the two traits, but it first has to be rescaled to get the genetic covariance:
  $\mathrm{Cov}(g_1,g_2) = \text{slope} \times M / \sqrt{N_1N_2}$
- Then standardize that covariance to get the genetic correlation:
  $r_g = \frac{\mathrm{Cov}(g_1,g_2)}{h_{1} h_{2}}$
- Under random mating and i.i.d. SNP effects, this is often interpreted as overlap of additive effects; in Part 2 we will see how assortative mating can weaken that interpretation.
:::

## Local LD is also useful {.smaller}

Local LD can blur causal interpretation, but because it is mostly nearby, measurable, and predictable, we can often model it and use it.


- Nearby SNPs can **tag** causal variants, so we do not need to observe every causal SNP directly
- Local LD allows **imputation** of untyped variants from typed ones
- Tools like **LDpred** can adjust SNP weights to account for LD structure, improving polygenic prediction
- **LD-score regression** exploits link between high LD and inflated test statistics to estimate heritability
- More broadly, local LD is often a **manageable nuisance**, not an arbitrary source of noise

## Summary: local LD both confounds and helps {.smaller}

::: {style="font-size: 0.8em;"}
| Step | Main lesson |
|---|----|
| Nearby SNPs correlate | Recombination makes nearby loci travel together |
| Marginal coefficients absorb tagged effects | So $\tilde{\beta}_j$ need not equal $\beta_j$ |
| Local LD is also useful | Tagging, imputation, and LD-aware methods exploit nearby structure |
| LD scores quantify local tagging | They summarize how much nearby variation a SNP captures |
| LD score regression exploits that structure | The same LD that complicates interpretation can be modeled |
:::

<br>

::: {.callout-note}
## Reminder:
$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
& + &
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)} \\
& &
\scriptstyle {\text{True Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$
:::

# Part 2: Assortative Mating and Long-Range LD {background-color="#2c3e50" style="color: white;"}

## What is assortative mating?

::: {.callout-warning}
## Partner similarity, $\text{Cov}(y_m, y_p) > 0$, can arise for different reasons. 

- **Assortative mating**: matching induces partner covariance
- **Convergence**: partners become more similar after pairing
- **Stratification**: partners are similar because both are shaped by the same background factors

Here, we assume that partner covariance is due to assortative mating, and we focus on the genetic consequences of that process.
:::

**Assortative mating (AM)** means that the **matching process itself** induces covariance between partners, rather than that covariance reflecting a shared cause. 


::: {.callout-note}
## Disassortative mating?
Technically, **assortative mating can be negative:** ($\text{Cov}(y_m, y_p) < 0$). However, this is rare in human populations. We therefore focus on positive assortative mating. 

The consequences of negative assortative mating are generally the same, but with opposite signs.
:::
## Consequences of assortative mating

**Key intuition:** Assortative mating induces correlations among causes by matching on the effects.

<br>

**Ordinary family resemblance**: <br> Covariance starts in their **genetic values** (and other causes) and is carried forward into their phenotypes 

::: {.callout-note appearance="minimal"}
Correlation flows from $g \to y$.
:::

<br>

**Assortative mating**: <br>
Covariance starts in their **phenotypes**, and is carried back into their genetic values.

::: {.callout-note appearance="minimal"}
Correlation flows from $y \to g$.
:::


Assortative mating therefore induces genetic similarity between partners.



## Predicted genetic value given phenotype
To quantify the induced genetic similarity, we **start at the phenotype** and ask: <br>
if we observe someone's phenotype $y$, what genetic value $g$ do we expect?

Assume centered variables and the simple additive model:

$$
y_i = g_i + e_i;
\quad
\small \text{Cov}(g,e)=0;
\quad
E[g_i] = E[e_i] = 0
\quad
\Rightarrow \quad
\text{Cov}(g,y)
= \text{Var}(g)
= \sigma_g^2
$$

Here, the regression slope of $g$ on $y$ is

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2}{\sigma_y^2}
= h^2
$$

Meaning the corresponding expected genetic value given phenotype is

$$
E[g_i \mid y_i] = h^2 y_i
$$


::: {.callout-warning}
## $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope even if $\text{Cov}(g,e) \neq 0$, but it no longer equals $h^2$.
<details>
<summary>Show the algebra</summary>
If $\text{Cov}(g,e)\neq 0$, the regression coefficient $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope, but that coefficient is no longer exactly narrow-sense heritability:

If $y=g+e$, then

$$
\text{Cov}(g,y)
= \text{Cov}(g, g+e)
= \text{Var}(g) + \text{Cov}(g,e)
$$

and

$$
\text{Var}(y)
= \text{Var}(g+e)
= \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

Hence

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2 + \sigma_{ge}}{\sigma_g^2 + \sigma_e^2 + 2\sigma_{ge}}
= h^2 + \frac{\sigma_{ge}}{\sigma_y^2}
$$

where $\sigma_{ge} = \text{Cov}(g,e)$ and $h^2 = \sigma_g^2/\sigma_y^2$.

So the regression coefficient equals $h^2$ only when $\text{Cov}(g,e)=0$.
If $\text{Cov}(g,e) > 0$, it is larger than $h^2$; if $\text{Cov}(g,e) < 0$, it is smaller.

</details>
:::

## Quantifying genetic similarity 

Applying $E[g \mid y] = h^2 y$ to both mates, the induced covariance between their genetic values is:

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}(h^2 y_m, h^2 y_p)
=
h^4 \text{Cov}(y_m,y_p)
$$

So the induced genotypic correlation between mates is

$$
r_g^{\mathrm{mates}}
=
\frac{(\sigma_g^2/\sigma_y^2)^2 \text{Cov}(y_m,y_p)}{\sigma_g^2}
=
\frac{\sigma_g^2}{\sigma_y^2}
\frac{\text{Cov}(y_m,y_p)}{\sigma_y^2}
= h^2 r_y^{\mathrm{mates}}
$$

::: {.callout-note}
The $h^4$ appears on the covariance scale because the phenotype-to-genotype regression slope ($\sigma_g^2/\sigma_y^2 = h^2$) is applied **once for each mate**. When we convert covariance to correlation, one power of $\sigma_g^2$ is divided back out, leaving the simpler result $r_g^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$. 
:::


## Direct versus indirect assortative mating {.smaller}

The preceding (and following) derivations assume that partners match directly on the **focal phenotype** $y$. 

In reality, they may match on a related signal or bundle of traits $s$ that correlates with $y$. 

::: {style="font-size: 0.8em;"}
| Case | What partners match on | Induced genotypic correlation between partners |
|--|----|---|
| **Direct assortment** | The focal phenotype itself: $s = y$ | $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,y)^2 r_y^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$ |
| **Indirect assortment** | A related signal or bundle of traits: $s \neq y$, but $\mathrm{Corr}(s,y) \neq 0$ |  $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,s)^2 r_s^{\mathrm{mates}}$ |
:::


::: {.callout-note}
## Key point
The genotypic correlation between partners depends on **how strongly the genetic value for the focal phenotype correlates with $s$**. Indirect assortative mating can imply **more or less** genetic resemblance between partners than you would otherwise expect.

In this lecture, we assume direct assortment where $s=y$, but note that assortment is indirect for many social traits (e.g., educational attainment).

:::

::: {.callout-tip}
You can think of $s$ as a secondary phenotype $y_2$, where genetic similarity depends on the genetic correlation between $y$ and $y_2$, the heritability of $y_2$, and the phenotypic partner correlation for $y_2$. Gene-environment correlations complicate things somewhat, but the same logic applies.
:::

## Genetic similarity spreads across loci

Rememeber, the genetic value is a weighted sum of SNPs: $g_i = \sum_{j=1}^{M} \beta_j x_{ij}$. 

Covariance between partners' genetic values is therefore the sum of all SNP-level covariances:

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}\!\left(\underbrace{\sum_{j=1}^{M} \beta_j x_{mj}}_{g_m}\ ,\  \underbrace{\sum_{k=1}^{M} \beta_k x_{pk}}_{g_p}\right)
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
 \beta_j\beta_k \,\text{Cov}(x_{mj}, x_{pk})
$$

Importantly, all these covariances **have the same sign** (i.e., they are all positive) for trait-associated SNPs, because the matching process aligns the genetic values of the partners.

::: {.callout-note}
## SNP-level covariance

To calculate $\text{Cov}(x_{mj}, x_{pk})$, we can use the same regression logic as before where we regress the two SNPs on their phenotype and combine that with the phenotypic covariance: $\frac{\text{Cov}(x_{mj}, y_m)}{\text{Var}(y_m)} \text{Cov}(y_m, y_p) \frac{\text{Cov}(y_p, x_{pk})}{\text{Var}(y_p)}$.
:::


## Why is this a problem? {.huge background-color="#5b0f0f" color="white" .center}
::: {.callout-important  appearance="simple"}
Once partner genomes become correlated, transmission and recombination **embeds the correlation structure** in the next following generation. 

Assortative mating across generations **induces a signed correlation structure between trait-associated alleles** (i.e., linkage disequilibrium), and that correlation structure can increase marginal SNP effects and family resemblance.

Now we will unpack how that happens.
:::

## What happens after *one* generation of AM 

First, we define the genetic values separately for the maternal and paternal haplotypes:

$$
\mcol{g_c^{m}} = \sum_{j=1}^{M} \beta_j \mcol{x_j^{m}}, 
\qquad
\pcol{g_c^{p}} = \sum_{j=1}^{M} \beta_j \pcol{x_j^{p}},
\qquad
g_c = \mcol{g_c^{m}} + \pcol{g_c^{p}}
$$

where $\mcol{x_j^{m}}$ and $\pcol{x_j^{p}}$ are the maternally and paternally transmitted alleles at SNP $j$, respectively. 

After only one generation of assortment, the transmitted alleles are still independent **within** haplotypes (assuming no relevant local LD):

$$
\operatorname{Cov}(\mcol{x_j^{m}}, \mcol{x_k^{m}}) = 0,
\qquad
\operatorname{Cov}(\pcol{x_j^{p}}, \pcol{x_k^{p}}) = 0
\qquad \text{for } j \neq k
$$

However, the transmitted alleles are correlated **across** haplotypes, because the parental genomes are correlated: 

$$
\text{Cov}(\mcol{g_c^{m}}, \pcol{g_c^{p}}) > 0
\qquad
\Rightarrow
\qquad
\text{Cov}(\mcol{x_j^{m}}, \pcol{x_k^{p}}) > 0 \small \text{ for trait-associated } j,k
$$


## Genetic variance after one generation of AM
Because the transmitted alleles are still independent within haplotypes, the variance of each haplotype contribution is just a sum of single-locus terms:
$$
\text{Var}(\mcol{g_c^{m}}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(\mcol{x_j^{m}}),
\qquad
\text{Var}(\pcol{g_c^{p}}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(\pcol{x_j^{p}})
$$

However, because the transmitted alleles from either parent are correlated, the total variance of the genetic value includes a positive covariance term:

$$
\text{Var}(g_c)
=
\text{Var}(\mcol{g_c^{m}})
+ \text{Var}(\pcol{g_c^{p}})
+ 2\text{Cov}(\mcol{g_c^{m}}, \pcol{g_c^{p}})
$$


::: {.callout-note}
## Terminology: trans vs cis covariance
Covariance **across** the maternal and paternal haplotypes is called **trans** covariance.

Covariance **within** the maternal or paternal haplotype is called **cis** covariance.
:::

## What happens after *several* generations of AM {.smaller}

Repeated assortment plus recombination gradually converts part of the inherited **trans** covariance into **cis** covariance. Written out fully, the three variance components become:

$$
{\scriptsize
\text{Var}(g_c)
=
\underbrace{
\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\mcol{x_j^{m}}, \mcol{x_k^{m}})
}_{\text{Var}\left(\mcol{g_c^{m}}\right)\ \text{(cis)}}
+
\underbrace{
\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\pcol{x_j^{p}}, \pcol{x_k^{p}})
}_{\text{Var}\left(\pcol{g_c^{p}}\right)\ \text{(cis)}}
+
\underbrace{
2\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\mcol{x_j^{m}}, \pcol{x_k^{p}})
}_{2\text{Cov}\left(\mcol{g_c^{m}}, \pcol{g_c^{p}}\right)\ \text{(trans)}}
}
$$

<br>

::: {style="font-size: 0.8em;"}
| When | Where | Interpretation |
|--|---|-----|
| 1 generation | 	Across haplotypes (trans)  | Maternal and paternal contributions become aligned |
| 2+ generations | Within haplotypes (cis) also builds up | Aligned alleles begin to co-occur within haplotypes |
:::

## Genetic variance across generations {.smaller}

Genetic variance increases across generations to an equilibrium::


::: {.columns  }
::: {.column width="55%"}
```{r}
#| label: s2-am-variance-trajectory
#| echo: false
#| cache: true
#| fig-align: center
#| fig-width: 6
#| fig-height: 5
#| fig-cap: ""
sg2_0 = 1
VE = 1
ry = 0.5
generations = 0:12

sg2 = numeric(length(generations))
sg2[1] = sg2_0
for (i in seq_len(length(generations) - 1)) {
  rg_mates = ry * sg2[i] / (sg2[i] + VE)
  sg2[i + 1] = sg2_0 / 2 + sg2[i] / 2 * (1 + rg_mates)
}

sg2_eq = uniroot(
  function(v) v - (sg2_0 / 2 + v / 2 * (1 + ry * v / (v + VE))),
  lower = sg2_0,
  upper = 10 * sg2_0
)$root

library(ggplot2)

df_var = data.frame(
  generation = generations,
  sg2 = sg2,
  baseline = sg2_0
)

ggplot(df_var, aes(x = generation, y = sg2)) +
  geom_ribbon(aes(ymin = baseline, ymax = sg2), fill = "#5b0f0f") +
  geom_ribbon(aes(ymin = 0, ymax = baseline), fill = "#cecece") +
  geom_line(color = "black", linewidth = 1.2) +
  geom_hline(aes(yintercept = sg2_0), color = "black",
             linewidth = 1, linetype = "dashed") +
  geom_hline(aes(yintercept = sg2_eq), color = "black",
             linewidth = 1, linetype = "dotted") +
  annotate(
    "text",
    x = 7.2,
    y = sg2_0 + 0.25 * (sg2_eq - sg2_0),
    label = "Added by LD between causal SNPs",
    color = "#ffffff",
    size = 6
  ) +
  scale_x_continuous(
    limits = c(0, max(generations)),
    breaks = generations
  ) +
  scale_y_continuous(
    breaks = seq(0,2,.25)
  ) +
  coord_cartesian(ylim = c(0, 1.08 * sg2_eq), clip = "off", expand = F) +
  labs(
    x = "Generations of assortment: t",
    caption = expression(h[0]^2 == 0.5 ~ ":" ~ r[y] == 0.5),
    y = expression(Genetic~variance:~sigma[g]^2)
  ) +
  theme_classic(base_size = 15) +
  theme(
    text = element_text(color = "black"),
    legend.position = "none" )
```

:::
::: {.column width="40%"}

::: {.callout-important}
## Heritability changes too
Changes in genetic variance changes the ratio of genetic to phenotypic variance: 
$$h_\infty^2 = \sigma_{g,\infty}^2 / (\sigma_{g,\infty}^2 + \sigma_{e}^2)$$

**Assortative mating therefore changes the heritability**. 

It typically **increases**, but with environmental transmission, the environmental variance changes too and the net effects are more complex:
$$h_\infty^2 = \sigma_{g,\infty}^2 / (\sigma_{g,\infty}^2 + \sigma_{e,\infty}^2 + 2\sigma_{g,e,\infty})$$

:::

:::
:::

::: {style="font-size: 0.7em;"}
<details>
<summary>Show recurrence equation for genetic variance</summary>

$$
\sigma^2_{g,t+1}
=
\underbrace{\frac{\sigma^2_{g,0}}{2}}_{\text{within-family}}
+ \quad
\underbrace{\sigma^2_{g,t}\frac{1 + r_{g,t}^{\mathrm{mates}}}{2}}_{\text{between-family}}
\qquad
\Rightarrow
\qquad
\sigma^2_{g,\infty}
=
\frac{\sigma^2_{g,0}}{1-r_{g,\infty}^{\mathrm{mates}}}
$$

</details>
:::


## Why should we care? {.huge background-color="#5b0f0f" color="white" .center}

<br>

::: {style="font-size: 1.5em;"}
Assortative mating in earlier generations means that **trait-associated SNPs are not independent from each other**.
:::

## Which estimands change once genotype covariance changes? {.smaller}

If assortative mating changes $\text{Cov}(x_j, x_k)$, then several things can change at once:

::: {style="font-size: 0.8em;"}
- **GWAS marginal effects**: $\tilde{\beta}_j$ absorbs long-range tagged signal
- **LDSC $h^2$**: the summary statistics and local LD scores no longer line up as cleanly as under random mating
- **GREML $h^2$**: Session 1's $\sigma_g^2 \mathbf{G}$ model do not capture the AM-induced covariance
- **Family resemblance**: relatives share the AM-induced covariance structure, so resemblance increases
- **Cross-trait $r_g$**: score correlation can rise without more effect correlation
:::

## Consequence 1: Marginal associations are inflated {.smaller}


::: {.callout-note}
## Reminder: the marginal association at SNP $j$ is

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
& + &
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)} \\
& &
\scriptstyle {\text{True Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

:::

Again set the rGE term aside and focus on the LD component. The sum is typically dominated by nearby loci (local LD). Under assortative mating, some of those covariance terms involve **distant** causal variants.

::: {.callout-important}
## The upward inflation is *specific to causal loci* (and loci that tag causal loci through local LD).

For these loci, the marginal association at SNP $j$ partly reflects genome-wide covariance structure, not only local biology.
:::

Because assortative mating changes marginal associations estimated by GWAS, downstream methods using GWAS summary statistics are affected.


## Consequence 2: LDSC can overstate $h^2$ {.smaller}

In Part 1, LDSC exploited that the expected $\chi^2$ statistic at SNP $j$ is proportional to heritability and its LD score, $\ell_j$:

$$
E[\chi_j^2 \mid \ell_j] = 1 + \frac{N h^2}{M}\ell_j + c
$$


- Under assortative mating, $\chi_j^2$ can be inflated by **long-range LD** between causal loci
- $\ell_j$ only capture **local LD** around SNP $j$

::: {.callout-important}
Because the causal SNPs are in LD *with each other*, all of their marginal effects are inflated by the same long-range covariance structure. Genetic effects are therefore effectively double-counted, and $h^2$ is overestimated.
:::

## Consequence 3: GREML behaves differently {.smaller}

In Session 1, GREML used the genomic relatedness matrix $\mathbf{G}$ to capture genetic similarity among unrelated people.

::: {style="font-size: 0.8em;"}
- $\mathbf{G}$ is built from **same-SNP, cross-person** similarity
- Assortative mating mainly adds **long-range LD between different SNPs**
- AM-induced extra genetic variance is not well captured by the GRM used by GREML
:::


GREML keeps fitting the Session 1 model $\sigma_g^2 \mathbf{G} + \sigma_e^2 \mathbf{I}$.

Under assortative mating, this is the wrong model and estimates will be biased.


::: {.callout-important}
- In realistic samples, GREML is generally **upwardly biased** under assortative mating
  - Unfortunately, the size of that bias can be hard to quantify and is highly dependent on sample size
- As sample sizes increase, the estimate drifts toward the random-mating heritability, $h^2_0$
  - In very large samples (N > 200K), GREML can therefore be **downwardly biased** because $h_0^2 < h_\infty^2$

:::

## Consequence 4: family resemblance changes too {.smaller}

At equilibrium, the genotypic correlation between $d$-degree relatives becomes

$$
r_{g,d} = \left(\frac{1 + r_g^{\mathrm{mates}}}{2}\right)^d
$$

Family-based estimators typically assume $r_{g,d} = (1/2)^d$, but that is only true under random mating

::: {.callout-note}
The exact bias is method dependent, but classic twin methods and midparent-offspring regression are **downwardly biased** under positive assortative mating.
:::


## Consequence 5: Cross-trait assortment affects $r_g$ {.smaller}

::: {.callout-tip}
## Reminder: two definitions of genetic correlation

$$
\text{Score correlation}\ = \text{Corr}(g_1, g_2),
\qquad
\text{Effect correlation}\ = \text{Corr}(\boldsymbol{\beta}_1, \boldsymbol{\beta}_2)
$$


If SNPs were i.i.d., these would be equal. Assortative mating can break this equivalence.

:::


**Cross-trait assortative mating (xAM)** means that people sort across two traits, or on a bundle of traits related to both $y_1$ and $y_2$.

Across generations, this **builds long-range LD between trait-increasing alleles across $y_1$ and $y_2$**.

This can change the score correlation even when effect correlation is zero

::: {.callout-important}
Estimators of $r_g$ (e.g., LDSC) are biased under xAM.

Notably, the resulting estimates do not correspond to either the *score correlation* or the *effect correlation*.
:::



## Summary: <br> assortative mating changes targets and estimators

::: {style="font-size: 0.8em;"}
| Step | Main lesson |
|--|-----|
| AM induces genetic similarity | $\text{Cov}(y_m,y_p) > 0 \quad \Rightarrow \quad \text{Cov}(g_m,g_p) > 0$ |
| Repeated AM | Long-range LD builds up over generations |
| Population target | Heritability itself can change because long-range LD adds genetic variance |
| GWAS / LDSC | Marginal associations absorb distant tagged signal, pushing summary-statistic $h^2$ upward |
| GREML | In realistic sample sizes, heritability estimates are usually upwardly biased, but converges to random-mating heritability as sample size increases |
| Relatives / family methods | Relatives become more genetically similar, so family-based estimators are affected too; the bias is method dependent |
| Cross-trait AM | Score correlation can exceed effect correlation under xAM. Estimators of $r_g$ are biased |
:::


# Part 3: Gene-Environment Correlations {background-color="#2c3e50" style="color: white;"}

## Same equation, now isolate the rGE term

rGE means genes and environments co-vary.
If $\mathrm{Cov}(x_j,e)\neq 0$, marginal SNP associations mix direct effects and rGE.

::: {.callout-note}
## Reminder: the marginal association at SNP $j$ is

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
& + &
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)} \\
& &
\scriptstyle {\text{True Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

:::

In Part 1 and Part 2, we focused on the LD term.

In Part 3, we keep that in the background and isolate the third term:

$$
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}.
$$

This is where gene-environment correlation enters the marginal SNP association.

## Where can $\text{Cov}(x_j,e)$ come from?
::: {style="font-size: 0.8em;"}
| Source | Why genotype and environment become correlated |
|--|----|
| Indirect genetic effects | Related people's genotypes shape the focal person's environment |
| Stratification / population structure | Allele frequencies and environments differ across groups |
| Residual confounding | Selection, ascertainment, or measurement structure |
| Assortative mating | Can increase and even induce covariance between inherited alleles and family environments |
:::

Different mechanisms can produce the **same statistical symptom**: $\text{Cov}(x_j,e) \neq 0$.
Across many loci, this can aggregate into $\text{Cov}(g,e) \neq 0$.

::: {.callout-important}
## Only passive gene-environment correlation is confounding
- **Passive $r_{ge}$**: Genotype and environment are correlated because of shared causes (e.g., parents)
- **Active / Evocative $r_{ge}$**: Genotype and environment are correlated because the focal person selects, creates, or evokes their environment based on their genotype (e.g., peers, partners, neighborhoods)
Crucially, **active and evocative gene-environment correlations are not confounding**. They are part of the causal pathway from genotype to phenotype, and are therefore mediators of the genetic effect. This is also important for how "direct" genetic effects are interpreted.
:::

## Why relatives' environmental effects induce $\mathrm{Cov}(g,e)$ {.smaller}

Write the phenotype as

$$
y_i = g_i + e_i,\qquad
e_i = e_i^{I} + e_i^*,\qquad
e_i^{I} = \sum_{k=1}^{M}\gamma_k x_{i\ell k}.
$$

Here $e_i^{I}$ is the part of the focal person's environment created by relatives.

For a specific relative $\ell$, shared inheritance implies a same-locus correlation
$\mathrm{Cov}(x_{ij},x_{i\ell j}) \neq 0$.
Therefore

$$
\mathrm{Cov}(x_{ij},e_i^{I})
=
\sum_{k=1}^{M}\gamma_k\,\mathrm{Cov}(x_{ij},x_{i\ell k})
\neq 0.
$$

At minimum, the $k=j$ term contributes; with long-range covariance structure (for example from AM), some $k\neq j$ terms can also contribute.

Aggregating over loci with $g_i = \sum_{j=1}^{M}\beta_j x_{ij}$:

$$
\mathrm{Cov}(g_i,e_i^{I})
=
\sum_{j=1}^{M}\sum_{k=1}^{M}\beta_j\gamma_k\,\mathrm{Cov}(x_{ij},x_{i\ell k}).
$$

This is the direct route from relatives' environmental effects to nonzero $\mathrm{Cov}(g,e)$.

## Consequence 1: marginal SNP associations include rGE

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
& + &
\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)} \\
& &
\scriptstyle {\text{True Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

If $\mathrm{Cov}(x_j,e)\neq 0$, the marginal SNP association absorbs rGE and is no longer a pure direct effect.

## Consequence 2: $\text{Cov}(g,y)$ is also affected

$$
y = g + e
\qquad \Longrightarrow \qquad
\text{Cov}(g,y)
=
\text{Cov}(g,g+e)
=
\text{Var}(g) + \text{Cov}(g,e)
$$

So with rGE, $\text{Cov}(g,y)$ no longer equals $\text{Var}(g)$.

## Consequence 3: variance decomposition becomes awkward

Keep these quantities conceptually separate:

$$
\text{Var}(g) \qquad \text{Var}(e) \qquad \text{Cov}(g,e) \qquad \text{Var}(y)
$$

With $y=g+e$:

$$
\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

The sign of the covariance matters:

| Sign of $\text{Cov}(g,e)$ | Interpretation |
|---|---|
| Positive | Genetically higher-valued individuals tend to experience environments that push in the same direction |
| Negative | Genetically higher-valued individuals tend to experience environments that push in the opposite direction |

::: {.callout-important}
$\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)$ is a **statistical identity**, not a clean causal partition of phenotypic variation.
:::

## Mini example: negative rGE

Take:

$$
\text{Var}(g)=0.40,\quad
\text{Var}(e)=0.80,\quad
\text{Cov}(g,e)=-0.10.
$$

This can happen when environments partially compensate for genetic liability.

Then:

$$
\text{Var}(y)=0.40+0.80+2(-0.10)=1.00
$$

$$
\text{Cov}(g,y)=\text{Var}(g)+\text{Cov}(g,e)=0.40-0.10=0.30
$$

Without rGE (same $\text{Var}(g)$ and $\text{Var}(e)$), we would instead have:

$$
\text{Var}(y)=1.20,\qquad
\text{Cov}(g,y)=0.40.
$$

Negative rGE therefore moves both quantities downward.

## Take-home map and bridge

::: {style="font-size: 0.86em;"}
| Quantity | What it means | What it does **not** automatically mean |
|---|---|---|
| $\text{Cov}(x_j,e)$ | SNP-level genotype-environment alignment | A direct SNP effect |
| $\tilde{\beta}_j$ | Marginal SNP association | A pure direct effect |
| $\text{Cov}(g,y)$ | Genetic value-phenotype association | Automatically equal to $\text{Var}(g)$ (equality only if $\text{Cov}(g,e)=0$) |
| $\text{Cov}(g,e)$ | Trait-level genetic-environmental alignment | A causal variance component by itself |
:::

::: {.callout-warning}
**Caution:** `rGE` and `GxE` are not the same thing. `rGE` means genes and environments are correlated (for example, $\text{Cov}(g,e) \neq 0$), while `GxE` means genetic effects change across environments (an interaction).
:::

::: {.callout-note}
Part 4 follows naturally: family designs change the comparison (toward within-family variation), which helps with several between-family rGE sources but does not remove all covariance structure.
:::

# Part 4: Family-based designs (overview) {background-color="#2c3e50" style="color: white;"}

## Why bring up families at all?

Much of this lecture has been about one central problem: the **population** coefficient $\tilde{\beta}_j$ is not always the same as the direct effect $\beta_j$.

Across unrelated people, SNP associations can reflect:

- local LD
- assortative mating
- stratification
- parental genetic nurture and other shared family environments

One could, in principle, try to adjust for all of these things by adding more and more covariates. But that is unrealistic, even for observed confounders like other SNPS let lonen unobserved confounders.

::: {.callout-note}
Family-based designs matter because they change the comparison. Instead of asking why people from different families differ, they compare relatives who share much of their background.
:::

## Core intuition

Within a family, many background features are held much more constant:

- the parents are the same
- ancestry background is much more similar
- much of the rearing environment is shared

What still differs is the random segregation of alleles from parents to offspring.

Recall from Session 1:

$$
x_{jc} =
\underbrace{\left(\frac{1}{2}x_{jp} + \epsilon_{jp}\right)}_{\pcol{x_j^{p}}}
+
\underbrace{\left(\frac{1}{2}x_{jm} + \epsilon_{jm}\right)}_{\mcol{x_j^{m}}}
$$

$$
\text{Var}(x_{jc})
= \frac{1}{4}\text{Var}(x_{jp}) + \frac{1}{4}\text{Var}(x_{jm}) + \frac{1}{2}\text{Cov}(x_{jp}, x_{jm})
+ \underbrace{\text{Var}(\epsilon_{jp} + \epsilon_{jm})}_{\text{segregation variance}}
$$

::: {.callout-note}
This is the callback to Session 1: the **random segregation deviations** do the heavy lifting. The key term is the segregation variance, and it is **random**. It comes from which alleles happened to be transmitted, not from stable differences in ancestry, parents, or shared family background.
:::

::: {.callout-note}
Across siblings, the parents are the same, so much of the family background is fixed. What differs is the realized draw of the $\epsilon$ terms. That is why family designs use **within-family random segregation** to get closer to direct effects than ordinary population GWAS.
:::

## What family designs are good for

| Question | High-level answer |
|---|---|
| Why can they help with stratification? | Relatives are compared against each other rather than against people from different ancestry backgrounds |
| Why can they help with parental genetic nurture? | Siblings share the same parents, so much of the parental background is held fixed |
| Why can they help with assortative mating? | They hold the parental pair fixed, removing much of the between-family assortment signal |
| Why are they appealing conceptually? | They target the same confounding problem we have discussed all day, but by changing the design rather than adding more covariates |

## What they do **not** magically solve

Family-based estimates are still not automatically equal to the direct causal effect.

| Remaining limitation | Why it remains |
|---|---|
| Local and long-range LD | Relatives still inherit correlated variants, including background correlation induced by assortative mating |
| Indirect effects within families | One relative's genotype can still affect another relative's environment |
| Non-shared environments | Siblings do not experience identical environments |
| Lower precision | Within-family contrasts use less genetic variation than population GWAS |

::: {.callout-important}
The right takeaway is not "family GWAS is perfect." The right takeaway is that it removes much of the **between-family confounding** that worries us in population GWAS.
:::

## Cleaner weights are not the same as unconfounded scores

Family designs mainly help with the **estimated SNP weight** by making the marginal coefficient closer to a direct effect.

But many downstream uses depend on both the weight and the realized genotype:

$$
\mathrm{PGS}_i^{F} = \sum_j \hat{\tilde{\beta}}_j^{F} x_{ij}
$$

That does **not** mean the score itself is automatically unconfounded.

Two problems can still remain at the SNP level:

- $x_{ij}$ can still be correlated with other causal SNPs through local or long-range LD
- $x_{ij}$ can still be correlated with environments, so $\text{Cov}(x_{ij}, e_i)$ need not vanish

::: {.callout-important}
Even a perfectly measured genetic value $g$ would still be associated with the phenotype through both direct and indirect pathways. A polygenic score has the same problem, so better SNP weights alone do not make it unconfounded.
:::

## Why adjust for parental or sibling polygenic scores?

At the polygenic score level, the same segregation logic still applies. A focal person's score can be written as:

$$
\mathrm{PGS}_F = \frac{1}{2}\left(\mathrm{PGS}_p + \mathrm{PGS}_m\right) + \Delta_F
$$

where:

- $\frac{1}{2}(\mathrm{PGS}_p + \mathrm{PGS}_m)$ is the expected family background in the score
- $\Delta_F$ is the **random segregation deviation** in the score

If the focal phenotype depends both on the focal genotype and on family-level genetic background, then regressing $y_F$ on $\mathrm{PGS}_F$ alone will mix those two sources of variation.

::: {.callout-note}
So the problem is the same as before: the observed polygenic score contains both a **between-family component** and a **within-family segregation component**.
:::

## Why the adjustment helps

If we fit

$$
y_F = b_1 \mathrm{PGS}_F + b_2 \overline{\mathrm{PGS}}_{parent} + \varepsilon_F
\qquad \text{or} \qquad
y_F = b_1 \mathrm{PGS}_F + b_2 \mathrm{PGS}_{sib} + \varepsilon_F
$$

then the extra family score absorbs part of the shared family-genetic background.

That means the remaining identifying variation in $\mathrm{PGS}_F$ is closer to the segregation term $\Delta_F$:

- adjusting for **parental** scores targets the family background more directly
- adjusting for a **sibling** score uses another child of the same parents as a proxy for that shared background
- in both cases, the coefficient on $\mathrm{PGS}_F$ becomes more within-family in spirit

::: {.callout-important}
This is why these controls can help: they soak up the part of the focal score that is predictable from the family, leaving more of the signal that comes from the focal person's **random segregation draw**.
:::

::: {.callout-note}
But again, this is not magic. LD, residual $\text{Cov}(x,e)$, sibling indirect effects, and measurement error in the scores can still matter.
:::

## Mixed-model family designs

Sibling comparisons are one kind of family design. Another is to use relatives inside a **mixed model** to separate direct and indirect genetic pathways.

## Intuition: ordinary GREML vs M-GCTA

Ordinary GREML asks:

- Are focal people who are slightly more genetically similar also slightly more phenotypically similar?

M-GCTA asks:

- Are focal people whose **mothers** are slightly more genetically similar also slightly more phenotypically similar?

::: {.callout-note}
In both cases the logic is covariance matching. If a source of genetic similarity creates phenotype similarity, give the mixed model the corresponding covariance matrix.
:::

## The M-GCTA model

In Session 1, GREML used a GRM to model covariance from individuals' own genotypes:

$$
\mathbf{y}_{F} = \mathbf{g}_{F} + \mathbf{e}_{F},
\qquad
\mathbf{g}_{F} \sim \mathcal{N}(\mathbf{0},\; \sigma_F^2 \mathbf{G}_{F})
$$

Here $F$ just means the **focal phenotype**. It could be an offspring phenotype, a maternal phenotype, or any other phenotype measured in one family member.

If a relative's genotype also helps shape that focal phenotype, then part of what we are calling "environment" has its own genetic covariance structure:

$$
\mathbf{y}_{F} = \mathbf{g}_{F} + \mathbf{g}_{R \rightarrow F} + \mathbf{e}_{F}
$$

::: {.callout-note}
The logic is exactly the same as ordinary GREML: if a pathway creates phenotype covariance, give the model the covariance matrix for that pathway. The only difference is that we now include covariance from a **relative's** genotype as well as the focal person's own genotype.
:::

In today's practical, the focal phenotype is the offspring and the relevant relative is the mother. So M-GCTA extends the direct-only GREML model by adding:

- a focal direct component based on $\mathbf{G}_{F} = \mathbf{G}_{oo}$
- a maternal indirect component based on $\mathbf{G}_{R} = \mathbf{G}_{mm}$
- a covariance component for overlap between direct and maternal effects

::: {.callout-important}
Following Eilertsen et al. (2021), this logic is not inherently tied to an offspring phenotype. The framework is defined relative to a **focal phenotype** and can be applied to phenotypes measured in different family members.
:::

::: {.callout-important}
This is why M-GCTA belongs in the family-design toolbox: it uses known family roles and multiple related genomes to ask whether a relative's genotype shapes the focal outcome through the environment, rather than forcing all genotype-related covariance into the focal person's direct effect.
:::

## Bridge to tomorrow

For this session, the main message is just:

$$
\text{Population GWAS: across-family variation}
\qquad \longrightarrow \qquad
\text{Family designs: within-family variation}
$$

That shift in identifying variation is why family-based methods are such an important complement to standard GWAS.

::: {.callout-note}
We will leave the design details, estimators, and tradeoffs for the family-based session, and keep today's treatment at the conceptual level.
:::

# Part 5: Practical {background-color="#2c3e50" style="color: white;"}

## Practical 1: GREML and M-GCTA with simulated trios

Session 2 continues the same `practical/sim_genomic.R` (or `practical/sim_genomic.jl`) script used in Session 1.

Today's new extension is:

1. Start from the direct-only setup already covered in Session 1
2. Simulate an offspring phenotype with a maternal indirect genetic effect
3. Refit the wrong direct-only GREML model
4. Fit the maternal model (M-GCTA) and compare the variance components

::: {.callout-note}
Parts 1-4 of the script are the same setup from Session 1, so they are not repeated on the slides here. If needed, rerun them in the script to recreate `X_tilde`, `yo`, `G`, `Goo`, and the direct-only GREML fit.
:::

## Starting point from Session 1 {.smaller}

By this point, you should already have the direct-only objects from the Session 1 practical:

- standardized genotypes `X_tilde`
- offspring phenotype `yo`
- genomic relationship matrix `G`
- offspring block `Goo`
- direct-only GREML fit

The following slides focus only on the **new Session 2 extension**.

<details>
<summary>Show Session 1 code</summary>

:::: {.panel-tabset}
### R
```{r}
#| label: s2-session1-setup
#| echo: false
#| eval: true
#| cache: true
set.seed(080318)
library(gaston)

N = 3000
M = 12000
p = 0.5
Xf = matrix(rbinom(N * M, 2, p), N, M)
Xm = matrix(rbinom(N * M, 2, p), N, M)
Xo = matrix(
  rbinom(N * M, 1, as.vector(Xf) / 2) +
  rbinom(N * M, 1, as.vector(Xm) / 2), N, M)
X = rbind(Xf, Xm, Xo)

idf = 1:N
idm = (N + 1):(2 * N)
ido = (2 * N + 1):(3 * N)

p_est = colMeans(X) / 2
X_tilde = scale(X, 2 * p_est, sqrt(2 * p_est * (1 - p_est)))
s2_g = 0.5
s2_e = 0.5
b = rnorm(M, 0, sqrt(s2_g / M))
g = as.vector(X_tilde %*% b)
e = rnorm(3 * N, 0, sqrt(s2_e))
y = g + e

yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
m_po = lm(yo ~ yp)

G = tcrossprod(X_tilde) / M
Goo = G[ido, ido]
m_direct = lmm.aireml(yo, K = Goo, verbose = FALSE)
```

### Julia
```julia
#| echo: false
#| eval: false
using LinearAlgebra, Distributions, DataFrames, GLM
using GREMLModels, Random
Random.seed!(080318)

N, M = 3000, 12000
p = 0.5
Xf = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xm = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xo = rand.(Bernoulli.(Xf ./ 2)) .+ rand.(Bernoulli.(Xm ./ 2))
X = vcat(Xf, Xm, Xo)

idf = 1:N
idm = N+1:2N
ido = 2N+1:3N

p_est = vec(mean(X, dims = 1)) ./ 2
X_tilde = (Float64.(X) .- (2 .* p_est)') ./ sqrt.(2 .* p_est .* (1 .- p_est))'

sigma2_g, sigma2_e = 0.5, 0.5
b = randn(M) * sqrt(sigma2_g / M)
g = X_tilde * b
e = randn(3N) * sqrt(sigma2_e)
y = g + e

yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
df1 = DataFrame(yo = yo, yp = yp)
mpo = lm(@formula(yo ~ yp), df1)

G = (X_tilde * X_tilde') / M
Goo = G[ido, ido]
Id = Diagonal(ones(N))
r = [Goo, Id]
mdirect = fit(GREMLModel, @formula(yo ~ 1), df1, r; verbose = false)
```

::::

</details>

## Part 5: Simulate a maternal indirect genetic effect {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-maternal-sim
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-session1-setup"
Sg = rbind(
  c(0.2, 0.2),
  c(0.2, 0.5)
)
s2_e = 0.1
B = matrix(rnorm(2 * M), M, 2) %*% chol(Sg / M)
g_m = X_tilde %*% B[, 1]
g_d = X_tilde %*% B[, 2]
e_2 = g_m[idm] + rnorm(N, 0, sqrt(s2_e))
yo_2 = g_d[ido] + e_2
```

### Julia
```julia
Sigma_g = [0.2 0.2; 0.2 0.5]
sigma2_e = 0.1
C = cholesky(Symmetric(Sigma_g / M)).U
B = randn(M, 2) * C
g_m = X_tilde * B[:, 1]
g_d = X_tilde * B[:, 2]
e2 = g_m[idm] + randn(N) * sqrt(sigma2_e)
yo2 = g_d[ido] + e2
```

::::

## Part 6: Fit the wrong direct-only model {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-wrong-direct
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-maternal-sim"
m_direct_2 = lmm.aireml(yo_2, K = Goo, verbose = FALSE)
m_direct_2$tau
```

### Julia
```julia
df2 = DataFrame(y = yo2)
mdirect2 = fit(GREMLModel, @formula(y ~ 1), df2, r; verbose = false)
```

::::

## Part 7: Fit the maternal model (M-GCTA) {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-mgcta
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-wrong-direct"
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
m_maternal = lmm.aireml(yo_2, K = list(Gmm, Dom, Goo), verbose = FALSE)
m_maternal$tau
```

### Julia
```julia
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
r2 = [Gmm, Dom, Goo, Id]
mmaternal = fit(GREMLModel, @formula(y ~ 1), df2, r2; verbose = false)
```

::::

## Interpretation checkpoint

After running the simulation, check that you can answer these questions:

1. Does parent-offspring regression recover something close to the simulated `h^2 = 0.5` in the direct-only setting?
2. In the direct-only GREML model, what do the estimated variance components look like before and after adding the maternal indirect effect?
3. Once you fit the maternal model, do the variance components move closer to the data-generating values?
4. Why does the direct-only model misattribute part of the maternal effect?


## Backup (optional): direct-indirect covariance derivation {.smaller}

If

$$
g_i^{D} = \sum_{j=1}^{M}\beta_j x_{ij},
\qquad
g_i^{I} = \sum_{k=1}^{M}\gamma_k x_{i\ell k},
$$

then

$$
\mathrm{Cov}(g_i^{D},g_i^{I})
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\gamma_k\,\mathrm{Cov}(x_{ij},x_{i\ell k}).
$$

This is AM-agnostic: long-range LD, relatedness, and other genotype correlation structure all enter through $\mathrm{Cov}(x_{ij},x_{i\ell k})$.



