---
title: "Gene-Environment Correlations and Assortative Mating"
subtitle: "ESSGN 2026 Workshop"
author: "Espen Eilertsen & Hans Fredrik Sunde"
date: 2026-03-25
format:
  revealjs:
    embed-resources: true
    theme: simple
    slide-number: true
    center: false
    chalkboard: false
    smaller: true
    scrollable: true
    transition: fade
    fig-align: center
    width: 1280
    height: 720
    margin: 0.04
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

<!-- Edit maternal/paternal math colors here if needed. -->
<div style="height: 0; overflow: hidden;">
$$
\newcommand{\mcol}[1]{\color{red}{#1}}
\newcommand{\pcol}[1]{\color{darkblue}{#1}}
$$
</div>

## Before we begin {.smaller}

This lecture is mostly **theoretical**. The goal is to understand why our usual genetic methods can become hard to interpret when correlation structure enters the picture.

- We will focus on mechanisms that can distort marginal SNP associations and variance decompositions.
- So the main takeaway is not a ready-to-use practical method, but a clearer picture of **what can go wrong** and why.

**Quick notation reminder**

- $x_{ij}$: genotype of individual $i$ at SNP / locus $j$
- $x_j$: shorthand for the genotype at SNP $j$ when the individual index is implicit
- $\beta_j$: direct causal effect of SNP $j$
- $\tilde{\beta}_j$: marginal GWAS coefficient at SNP $j$
- $g_i = \sum_{j=1}^{M} \beta_j x_{ij}$: additive genetic value
- $e_i$: environmental / residual component
- $y_i = g_i + e_i$: phenotype
- $M$: number of SNPs / loci in the model


## What does a GWAS estimate?
The infinitesimal model says that the phenotype is a sum of many small genetic effects plus an environment:
$$
y_i = \sum_{j=1}^{M} \beta_j x_{ij} + e_i
$$

In a GWAS, each SNP $j$ is tested separately in a linear model:

$$
y_i = \alpha_j + \tilde{\beta}_j x_{ij} + \dots + \varepsilon_{ij}
$$


$\tilde{\beta}_j$ is the **marginal coefficient** measuring the ***association*** between $x_{ij}$ and $y_i$: 

$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\color{red}{\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}}
& + &
\color{blue}{\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}} \\
& &
\scriptstyle {\text{Causal Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$

::: {.callout-note appearance="minimal"}
Here we ignore sampling error, imperfect genetic overlap between GWAS and target phenotypes, and heterogeneity across GWAS samples
:::

## Lecture Outline

The unifying theme today is that different sources of confounding ultimately comes down to the same thing: **marginal associations can be confounded by correlation structure**.

1. **Part 1: Local linkage disequilibrium (LD)**

   Why nearby correlated SNPs both distort and inform marginal effect estimates
2. **Part 2: Assortative mating and long-range LD**

   How mating behavioirs affect correlations between distant loci
3. **Part 3: Gene-environment correlations**

   How indirect effects, stratification, and correlated environments enter SNP estimates and variance decompositions
4. **Part 4: Family-based designs (overview)**

   Why within-family comparisons are useful, and what they can and cannot solve
5. **Part 5: Practical**

   Practical 1: GREML and maternal indirect genetic effects with M-GCTA
   Practical 2: SNP heritability and a genetic correlation with LD score regression

# Part 1: Local Linkage Disequilibrium {background-color="#2c3e50" style="color: white;"}

## Nearby SNPs correlate

Recombination shuffles **chromosome segments**, not isolated SNPs. Loci that are close together are less likely to be separated by recombination and therefore tend to be inherited together. 

<table style="border-collapse: collapse; font-family: Lato;">
  <tr><td colspan="11" style="height:10px;"></td></tr>
   <tr>
    <td style="border: none; text-align: right;"><b>$t_{grandmother}$</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">0</td>
    <td style="background:#d6eaf8; border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">1</td>
  </tr>

  <tr>
    <td style="border: none; text-align: right;"><b>$t_{grandfather}$</b></td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="border:1px solid #ccc;">1</td>
    <td style="border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">1</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
    <td style="background:#fadbd8; border:1px solid #ccc;">0</td>
  </tr>

  <tr><td colspan="11" style="height:10px;"></td></tr>

  <tr>
    <td style="border: none; text-align: right;"><b>$t_{parent}$</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#d6eaf8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>1</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
    <td style="background:#fadbd8; border:1px solid #ccc;"><b>0</b></td>
  </tr>
   <tr><td colspan="11" style="height:10px;"></td></tr>

</table>


This makes nearby SNPs correlated in the population: $\text{Cov}(x_j, x_k) \neq 0$ for nearby $j,k$

This is called (local) **linkage disequilibrium (LD)**.

::: {.callout-note}
If two nearby SNPs are often inherited together, they are statistically *dependent*: once you know the allele at one SNP, you can make a good guess about the allele at the other.
:::


## A two-SNP toy example


Assume the true model is:

$$
y_i = \beta_1 x_{i1} + e_i
$$

where $x_1$ is causal and is in LD with $x_2$, a non-causal SNP:

$$
\beta_1 \neq 0, \qquad \beta_2 = 0, \qquad \text{Cov}(x_1,x_2) \neq 0
$$

A GWAS tests each SNP separately. For $x_2$, the marginal coefficient is:

$$
\tilde{\beta}_2
=
\color{grey}{\beta_2}
\color{grey}{+}
\beta_1 \frac{\text{Cov}(x_2,x_1)}{\text{Var}(x_2)}
$$

In other words, $x_2$ will be associated with the phenotype even though it has **no direct causal effect**.


## Generalizing to many SNPs

Recall that the marginal coefficient at SNP $j$ is (ignoring the rGE term for now):

$$
\tilde{\beta}_j = \beta_j
+ \sum_{k \ne j} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
$$

If we standardize the SNPs so that $\text{Var}(x_j)=1$, then $\text{Cov}(x_j,x_k)=r_{jk}$ and this becomes

$$
\tilde{\beta}_j = \sum_{k=1}^M r_{jk}\beta_k
$$

So the marginal coefficient at SNP $j$ is a weighted sum of the direct effects at correlated SNPs (including itself). A SNP that is correlated with **many** nearby variants can therefore pick up more tagged signal.



## We can exploit this structure {.huge background-color="#5b0f0f" color="white .center}

<br>

If a trait is heritable, then SNPs correlated with many nearby variants should have inflated test statistics on average, because they tag more causal variation in the region.

## What is an LD score?

For SNP $j$, the **LD score** is the sum of squared correlations with nearby SNPs:

$$
\ell_j = \sum_{k \in \mathcal{N}(j)} r_{jk}^2
$$

where $\mathcal{N}(j)$ is a local window around SNP $j$. 

::: {.callout-tip}
# Interpretation
A high LD score means the SNP is highly correlated with many nearby variants, and is more likely to tag causal variation in the region. 

A low LD score means the SNP is more isolated and less likely to tag nearby causal effects.
:::

## LD score regression

If SNPs with higher LD scores tag more nearby causal variation, then under polygenicity their test statistics should be larger on average. This relationship is proportional to the heritability. We can therefore exploit this structure to estimate SNP heritability from GWAS summary statistics, using a method called **LD score regression (LDSC)**.

$$
E[\chi_j^2 \mid \ell_j] = 1 + \frac{N h^2}{M}\ell_j + c
$$

where $\chi_j^2 = z_j^2$, and $z_j = \hat{\tilde{\beta}}_j/\text{SE}(\hat{\tilde{\beta}}_j)$ for SNP $j$.

::: {.callout-tip}
## High-level interpretation
- Regress GWAS test statistics on LD scores
- The **slope** tracks polygenic signal, but it must be rescaled to get heritability:
  $h^2 = \text{slope} \times M / N$
- The **intercept**, $c$, tracks confounding or other inflation

Other considerations include  appropriate regression weights to handle LD-dependent redundancy, plus a well-matched LD reference panel, careful GWAS QC, and checking for sample overlap or ancestry mismatch.
:::

## Genetic correlation from cross-trait LDSC {.smaller}

For two traits, the same idea extends to the product of the two marginal $z$-statistics:

$$
E[z_{1j} z_{2j} \mid \ell_j]
=
0 + 
r_g \,\ell_j
\left(
\frac{\sqrt{N_1 N_2}}{M} \, h_{1} h_{2}
\right)
+ c_{12}
$$

where $r_g$ is the **genetic correlation**, summarizing how strongly the two traits share genetic signal.

::: {.callout-tip}
## High-level interpretation
- Regress cross-trait $z$-score products on LD scores
- The slope tracks **shared genetic signal** across the two traits, but it first has to be rescaled to get the genetic covariance:
  $\mathrm{Cov}(g_1,g_2) = \text{slope} \times M / \sqrt{N_1N_2}$
- Then standardize that covariance to get the genetic correlation:
  $r_g = \frac{\mathrm{Cov}(g_1,g_2)}{h_{1} h_{2}}$
- Under random mating and i.i.d. SNP effects, this is often interpreted as overlap of additive effects; in Part 2 we will see how assortative mating can weaken that interpretation.
:::

## Summary: local LD both confounds and helps

| Step | Main lesson |
|---|---|
| Nearby SNPs correlate | Recombination makes nearby loci travel together |
| Marginal coefficients absorb tagged effects | So $\tilde{\beta}_j$ need not equal $\beta_j$ |
| LD scores quantify local tagging | They summarize how much nearby variation a SNP captures |
| LD score regression exploits that structure | The same LD that complicates interpretation can be modeled |

::: {.callout-note}
## Reminder:
$$
\begin{array}{r c c c c c c}
\tilde{\beta}_j
& = &
\beta_j
& + &
\color{red}{\sum_{k \ne j}^{M} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}}
& + &
\color{blue}{\frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}} \\
& &
\scriptstyle {\text{Causal Effect}}
& &
\scriptstyle {\text{Linkage Disequilibrium}}
& &
\scriptstyle {\text{rGE}}
\end{array}
$$
:::

# Part 2: Assortative Mating and Long-Range LD {background-color="#2c3e50" style="color: white;"}

## What is assortative mating?

Partner similarity, $\text{Cov}(y_m, y_p) > 0$, can arise for different reasons. 

**Assortative mating (AM)** means that the **matching process itself** induces covariance between partners, rather than that covariance reflecting a shared cause. 

::: {.callout-important}
## Observed partner covariance is a **symptom**, not yet an explanation.

- **Assortative mating**: matching induces partner covariance
- **Convergence**: partners become more similar after pairing
- **Stratification**: partners are similar because both are shaped by the same background factors

Here, we assume that partner covariance is due to assortative mating, and we focus on the genetic consequences of that process.
:::

::: {.callout-note}
## Negative assortative mating?
Technically, assortative mating can be negative if partners are more dissimilar than expected by chance. However, we focus on positive assortative mating here because it is much more common in human populations. The consequences of negative assortative mating are generally the same, but with opposite signs.
:::
## Consequences of assortative mating

::: {.callout-tip}
## Useful intuition
Assortative mating induces correlations among causes by matching on the effects. 
:::


**Ordinary family resemblance**: <br> Covariance starts in the **causes** and is diluted as it shows up in the phenotype 

::: {.callout-note appearance="minimal"}
Correlation flows from $g \to y$.
:::

**Assortative mating**: <br>
Covariance starts on the **phenotype**. If partners are phenotypically similar, then part of that covariance is carried back into the genetic values.

::: {.callout-note appearance="minimal"}
Correlation flows from $y \to g$.
:::


Assortative mating therefore induces genetic similarity between partners.



## Predicted genetic value given phenotype
To quantify the induced genetic similarity, we start at the **phenotype** and ask: if we observe someone's phenotype $y$, what genetic value $g$ do we expect?

Assume centered variables and the simple additive model:

$$
y_i = g_i + e_i;
\quad
\small \text{Cov}(g_i,e_i)=0;
\quad
E[g_i] = E[e_i] = 0
\quad
\Rightarrow \quad
\text{Cov}(g_i,y_i)
= \text{Var}(g_i)
= \sigma_g^2
$$

Here, the regression slope of $g$ on $y$ is

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2}{\sigma_y^2}
= h^2
$$

Meaning the corresponding expected genetic value given phenotype is

$$
E[g_i \mid y_i] = h^2 y_i
$$


::: {.callout-warning}
## $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope even if $\text{Cov}(g,e) \neq 0$, but it no longer equals $h^2$.
<details>
<summary>Show the algebra</summary>
If $\text{Cov}(g,e)\neq 0$, the regression coefficient $\text{Cov}(g,y)/\text{Var}(y)$ is still the relevant slope, but that coefficient is no longer exactly narrow-sense heritability:

If $y=g+e$, then

$$
\text{Cov}(g,y)
= \text{Cov}(g, g+e)
= \text{Var}(g) + \text{Cov}(g,e)
$$

and

$$
\text{Var}(y)
= \text{Var}(g+e)
= \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

Hence

$$
b_{g \mid y}
= \frac{\text{Cov}(g,y)}{\text{Var}(y)}
= \frac{\sigma_g^2 + \sigma_{ge}}{\sigma_g^2 + \sigma_e^2 + 2\sigma_{ge}}
= h^2 + \frac{\sigma_{ge}}{\sigma_y^2}
$$

where $\sigma_{ge} = \text{Cov}(g,e)$ and $h^2 = \sigma_g^2/\sigma_y^2$.

So the regression coefficient equals $h^2$ only when $\text{Cov}(g,e)=0$.
If $\text{Cov}(g,e) > 0$, it is larger than $h^2$; if $\text{Cov}(g,e) < 0$, it is smaller.

</details>
:::

## Quantifying genetic similarity 

Applying $E[g \mid y] = h^2 y$ to both mates, the induced covariance component is:

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}(h^2 y_m, h^2 y_p)
=
h^4 \text{Cov}(y_m,y_p)
$$

So the induced genotypic correlation between mates is

$$
r_g^{\mathrm{mates}}
=
\frac{(\sigma_g^2/\sigma_y^2)^2 \text{Cov}(y_m,y_p)}{\sigma_g^2}
=
\frac{\sigma_g^2}{\sigma_y^2}
\frac{\text{Cov}(y_m,y_p)}{\sigma_y^2}
= h^2 r_y^{\mathrm{mates}}
$$

::: {.callout-note}
The $h^4$ appears on the covariance scale because the phenotype-to-genotype regression slope ($\sigma_g^2/\sigma_y^2 = h^2$) is applied **once for each mate**. When we convert covariance to correlation, one power of $\sigma_g^2$ is divided back out, leaving the simpler result $r_g^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$. 
:::


## Direct versus indirect assortative mating {.smaller}

The preceding (and following) derivations assume that partners match directly on the **focal phenotype** $y$. 

In reality, they may match on a related signal or bundle of traits $s$ that correlates with $y$. We call $s$ the **sorting factor**.

::: {style="font-size: 0.8em;"}
| Case | What partners match on | Induced genotypic correlation between partners |
|--|----|---|
| **Direct assortment** | The focal phenotype itself: $s = y$ | $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,y)^2 r_y^{\mathrm{mates}} = h^2 r_y^{\mathrm{mates}}$ |
| **Indirect assortment** | A related signal or bundle of traits: $s \neq y$, but $\mathrm{Corr}(s,y) \neq 0$ |  $r_g^{\mathrm{mates}} = \mathrm{Corr}(g,s)^2 r_y^{\mathrm{mates}}$ |
:::


::: {.callout-note}
## Key point
The genotypic correlation between partners depends on **how strongly the sorting factor $s$ correlates with the genetic causes of the focal phenotype**. Indirect assortative mating can imply **more or less** genetic resemblance between partners than you would otherwise expect.

In this lecture, we assume direct assortment where $s=y$, but note that assortment is indirect for many social traits (e.g., educational attainment).
:::

## Genetic similarity spreads across many SNP pairs

The genetic value is a weighted sum of SNPs: $g_i = \sum_{j=1}^{M} \beta_j x_{ij}$. Genotypic covariance between mates therefore expands as

$$
\text{Cov}(g_m,g_p)
=
\text{Cov}\!\left(\underbrace{\sum_{j=1}^{M} \beta_j x_{mj}}_{g_m}\ ,\  \underbrace{\sum_{k=1}^{M} \beta_k x_{pk}}_{g_p}\right)
=
\sum_{j=1}^{M}\sum_{k=1}^{M}
 \beta_j\beta_k \,\text{Cov}(x_{mj}, x_{pk})
$$

Assortative mating therefore induces many small SNP-level correlations across partners. Importantly, all these correlations are **positive** for trait-associated SNPs, because the matching process aligns the genetic values of the partners.

::: {.callout-note}
To calculate $\text{Cov}(x_{mj}, x_{pk})$, we can use the same regression logic as before where we regress the two SNPs on their phenotype and combine that with the phenotypic covariance: $\frac{\text{Cov}(x_{mj}, y_m)}{\text{Var}(y_m)} \text{Cov}(y_m, y_p) \frac{\text{Cov}(y_p, x_{pk})}{\text{Var}(y_p)}$.
:::


## Why is this a problem? {.huge background-color="#5b0f0f" color="white" .center}
::: {.callout-important  appearance="simple"}
Once partner genomes become correlated, transmission and recombination **embeds the correlation structure** in the next following generation. 

Assortative mating across generations induces a signed correlation structure between trait-associated alleles (i.e., linkage disequilibrium), and that correlation structure can increase marginal SNP effects and family resemblance.

Now we will unpack how that happens.
:::

## What happens after *one* generation of AM 

First, we define the genetic values separately for the maternal and paternal haplotypes:

$$
\mcol{g_c^{m}} = \sum_{j=1}^{M} \beta_j \mcol{x_j^{m}}, 
\qquad
\pcol{g_c^{p}} = \sum_{j=1}^{M} \beta_j \pcol{x_j^{p}},
\qquad
g_c = \mcol{g_c^{m}} + \pcol{g_c^{p}}
$$

where $\mcol{x_j^{m}}$ and $\pcol{x_j^{p}}$ are the maternally and paternally transmitted alleles at SNP $j$, respectively. 

After only one generation of assortment, the transmitted alleles are still independent **within** haplotypes (assuming no relevant local LD):

$$
\operatorname{Cov}(\mcol{x_j^{m}}, \mcol{x_k^{m}}) = 0,
\qquad
\operatorname{Cov}(\pcol{x_j^{p}}, \pcol{x_k^{p}}) = 0
\qquad \text{for } j \neq k
$$

However, the transmitted alleles are correlated **across** haplotypes, because the parental genomes are correlated: 

$$
\text{Cov}(\mcol{g_c^{m}}, \pcol{g_c^{p}}) > 0
\qquad
\Rightarrow
\qquad
\text{Cov}(\mcol{x_j^{m}}, \pcol{x_k^{p}}) > 0 \small \text{ for trait-associated } j,k
$$


## Genetic variance after one generation of AM
Because the transmitted alleles are still independent within haplotypes, the variance of each haplotype contribution is just a sum of single-locus terms:
$$
\text{Var}(\mcol{g_c^{m}}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(\mcol{x_j^{m}}),
\qquad
\text{Var}(\pcol{g_c^{p}}) = \sum_{j=1}^{M} \beta_j^2 \text{Var}(\pcol{x_j^{p}})
$$

However, because the transmitted alleles from either parent are correlated, the total variance of the genetic value includes a positive covariance term:

$$
\text{Var}(g_c)
=
\text{Var}(\mcol{g_c^{m}})
+ \text{Var}(\pcol{g_c^{p}})
+ 2\text{Cov}(\mcol{g_c^{m}}, \pcol{g_c^{p}})
$$


::: {.callout-note}
## Terminology: trans vs cis covariance
Covariance **across** the maternal and paternal haplotypes is called **trans** covariance.

Covariance **within** the maternal or paternal haplotype is called **cis** covariance.
:::

## What happens after *several* generations of AM {.smaller}

Repeated assortment plus recombination gradually converts part of the inherited **trans** covariance into **cis** covariance. Written out fully, the three variance components become:

$$
{\scriptsize
\text{Var}(g_c)
=
\underbrace{
\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\mcol{x_j^{m}}, \mcol{x_k^{m}})
}_{\text{Var}\!\left(\mcol{g_c^{m}}\right)\ \text{(cis)}}
+
\underbrace{
\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\pcol{x_j^{p}}, \pcol{x_k^{p}})
}_{\text{Var}\!\left(\pcol{g_c^{p}}\right)\ \text{(cis)}}
+
\underbrace{
2\sum_{j=1}^{M}\sum_{k=1}^{M}
\beta_j\beta_k\,
\text{Cov}(\mcol{x_j^{m}}, \pcol{x_k^{p}})
}_{2\text{Cov}\!\left(\mcol{g_c^{m}}, \pcol{g_c^{p}}\right)\ \text{(trans)}}
}
$$

<br>

::: {style="font-size: 0.8em;"}
| When | Where | Interpretation |
|--|---|-----|
| 1 generation | 	Across haplotypes (trans)  | Maternal and paternal contributions become aligned |
| 2+ generations | Within haplotypes (cis) also builds up | Aligned alleles begin to co-occur within haplotypes |
:::

## Genetic variance across generations {.smaller}

The genetic variance increases asymptotically towards an equilibrium following the recurrence

$$
\sigma^2_{g,t+1}
=
\underbrace{\frac{\sigma^2_{g,0}}{2}}_{\text{within-family}}
+ \quad
\underbrace{\sigma^2_{g,t}\frac{1 + r_{g,t}^{\mathrm{mates}}}{2}}_{\text{between-family}}
\quad
\Rightarrow
\quad
\sigma^2_{g,\infty}
=
\frac{\sigma^2_{g,0}}{1-r_{g,\infty}^{\mathrm{mates}}}
$$

::: {.columns  }
::: {.column width="66%"}
```{r}
#| label: s2-am-variance-trajectory
#| echo: false
#| cache: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 5
#| fig-cap: ""
sg2_0 = 1
VE = 1
ry = 0.5
generations = 0:12

sg2 = numeric(length(generations))
sg2[1] = sg2_0
for (i in seq_len(length(generations) - 1)) {
  rg_mates = ry * sg2[i] / (sg2[i] + VE)
  sg2[i + 1] = sg2_0 / 2 + sg2[i] / 2 * (1 + rg_mates)
}

sg2_eq = uniroot(
  function(v) v - (sg2_0 / 2 + v / 2 * (1 + ry * v / (v + VE))),
  lower = sg2_0,
  upper = 10 * sg2_0
)$root

library(ggplot2)

df_var = data.frame(
  generation = generations,
  sg2 = sg2,
  baseline = sg2_0
)

ggplot(df_var, aes(x = generation, y = sg2)) +
  geom_ribbon(aes(ymin = baseline, ymax = sg2), fill = "#5b0f0f") +
  geom_ribbon(aes(ymin = 0, ymax = baseline), fill = "#cecece") +
  geom_line(color = "black", linewidth = 1.2) +
  geom_hline(aes(yintercept = sg2_0), color = "black",
             linewidth = 1, linetype = "dashed") +
  geom_hline(aes(yintercept = sg2_eq), color = "black",
             linewidth = 1, linetype = "dotted") +
  annotate(
    "text",
    x = 7.2,
    y = sg2_0 + 0.25 * (sg2_eq - sg2_0),
    label = "Added by LD between causal SNPs",
    color = "#ffffff",
    size = 8
  ) +
  scale_x_continuous(
    limits = c(0, max(generations)),
    breaks = generations
  ) +
  scale_y_continuous(
    breaks = seq(0,2,.25)
  ) +
  coord_cartesian(ylim = c(0, 1.08 * sg2_eq), clip = "off", expand = F) +
  labs(
    x = "Generations of assortment: (t)",
    caption = "Assuming baseline heritability of 50% and partner correlation of 0.5",
    y = expression(Genetic~variance:~sigma[g]^2)
  ) +
  theme_classic(base_size = 17) +
  theme(
    text = element_text(color = "black"),
    legend.position = "none" )
```

:::
::: {.column width="34%"}



::: {.callout-important}
## Heritability changes too
Changes in genetic variance changes the ratio of genetic to phenotypic variance: 
$$h_\infty^2 = \sigma_{g,\infty}^2 / (\sigma_{g,\infty}^2 + \sigma_{e,\infty}^2)$$

**Assortative mating therefore changes the heritability**. Usually, it **increases**, but with correlated environmental effects, the environmental variance changes too and the net effects are more complex:
$$h_\infty^2 = \sigma_{g,\infty}^2 / (\sigma_{g,\infty}^2 + \sigma_{e,\infty}^2 + 2\sigma_{g,e,\infty})$$

:::

:::
:::



## Why should we care? {.huge background-color="#5b0f0f" color="white" .center}

<br>

::: {style="font-size: 1.5em;"}
Assortative mating in earlier generations means that **trait-associated SNPs are not independent from each other**.
:::

## Which estimands change once genotype covariance changes? {.smaller}

If assortative mating changes $\text{Cov}(x_j, x_k)$, then two things can change at once:

::: {style="font-size: 0.8em;"}
- **GWAS marginal effects**: $\tilde{\beta}_j$ absorbs long-range tagged signal, so hits look less purely local
- **LDSC $h^2$**: the summary statistics and local LD scores no longer line up as cleanly as under random mating
- **REML / GREML $h^2$**: Session 1's $\sigma_g^2 \mathbf{G}$ model no longer spans all of the AM-induced covariance
- **Family resemblance**: relatives inherit the AM-induced covariance structure, so resemblance increases
- **Cross-trait $r_g$**: score correlation can rise without more effect correlation, so bivariate estimators can overstate overlap
:::

## Consequence 1: marginal SNP effects are no longer just local {.smaller}

Again set the rGE term aside and focus on the LD component:

$$
\tilde{\beta}_j
= \beta_j
+ \sum_{k \ne j} \beta_k \frac{\text{Cov}(x_j, x_k)}{\text{Var}(x_j)}
$$

Under local LD, the sum is dominated by nearby loci. Under assortative mating, some of those covariance terms can involve **distant** causal variants.

::: {.callout-note}
So the upward inflation is not generic to every SNP.

It is largest for SNPs that are themselves **causal**, or that sit close enough to causal loci to tag them through LD. For those loci, a marginal hit at SNP $j$ can partly reflect genome-wide covariance structure, not only local biology at SNP $j$.

That means the GWAS summary statistics are already altered before any downstream summary-statistic method is applied.
:::

## Consequence 2: LDSC can overstate $h^2$ {.smaller}

In Part 1, the LDSC logic was:

$$
E[\chi_j^2 \mid \ell_j] = 1 + \frac{N h^2}{M}\ell_j + c
$$

where $\ell_j$ summarizes **local** LD around SNP $j$.

- Under assortative mating, some of the extra signal in $\chi_j^2$ comes from **distant** trait-relevant loci
- LDSC still uses **local** LD scores, so that genome-wide dependence is not well represented in the regression
- The slope can therefore attribute too much signal to heritability, so LDSC can **overestimate $h^2$**

::: {.callout-important}
The key misspecification is upstream: assortative mating changes the **marginal coefficients themselves**, so LDSC starts from summary statistics that already mix direct effects with induced long-range covariance.

LDSC is therefore not just reading off the new genetic variance after assortative mating; it can overestimate even that.
:::

::: {.callout-note appearance="minimal"}
Source: Border et al. (2022, *Nature Communications*) and Supplementary Information.
:::

## Consequence 3: GREML behaves differently {.smaller}

In Session 1, GREML used the genomic relatedness matrix $\mathbf{G}$ to capture genetic similarity among unrelated people.

::: {style="font-size: 0.8em;"}
- $\mathbf{G}$ is built from **same-SNP, cross-person** similarity
- Assortative mating mainly adds **long-range LD between different SNPs**
- AM-induced extra genetic variance is not well captured by the GRM used by GREML
:::

::: {.callout-important}
GREML keeps fitting the Session 1 model $\sigma_g^2 \mathbf{G} + \sigma_e^2 \mathbf{I}$.

Under assortative mating, this is the wrong model and estimates will be biased.

- In realistic samples, GREML is generally **upwardly biased** under assortative mating
  - Unfortunately, the size of that bias can be hard to quantify and is highly dependent on sample size
- As sample sizes increase, the estimate drifts toward the random-mating heritability, $h^2_0$
  - In very large samples (N > 200K), GREML can therefore be **downwardly biased** because $h_0^2 < h_\infty^2$

:::

## Consequence 4: family resemblance changes too {.smaller}

At equilibrium, the genotypic correlation between $d$-degree relatives becomes

$$
r_{g,d} = \left(\frac{1 + r_g^{\mathrm{mates}}}{2}\right)^d
$$

Family-based estimators typically assume $r_{g,d} = (1/2)^d$, but that is only true under random mating

::: {.callout-note}
The exact bias is method dependent, but classic twin methods and midparent-offspring regression are **downwardly biased** under positive assortative mating.
:::


## Generalization: xAM can inflate $r_g$ {.smaller}

Session 1 distinguished two related ideas:

- *Score correlation*, $\text{Corr}(g_1, g_2)$
- *Effect correlation*, $\text{Corr}(\boldsymbol{\beta}_1, \boldsymbol{\beta}_2)$

::: {.callout-note}
If SNPs were i.i.d., these would line up. Part 2 has now given one concrete reason they can separate: assortative mating creates long-range LD across loci.
:::

If people sort on one trait, or on a bundle of traits related to both $y_1$ and $y_2$, then alleles increasing trait 1 can become correlated with alleles increasing trait 2 across the genome.

- Score correlation can therefore increase even when effect correlation does not
- So bivariate GWAS / LDSC can overstate biological overlap or pleiotropy when working from marginal summary statistics

::: {.callout-note appearance="minimal"}
Source: Border et al. (2022, *Science*) and Supplementary Information.
:::

## Summary: assortative mating changes targets and estimators

| Step | Main lesson |
|---|---|
| Mates match on phenotype | Start with $\text{Cov}(y_m,y_p) > 0$ |
| Repeated AM | Trans covariance appears first; cis covariance and long-range LD build up over generations |
| Population target | Heritability itself can change because long-range LD adds genetic variance |
| GWAS / LDSC | Marginal summary statistics absorb distant tagged signal, pushing summary-statistic $h^2$ upward |
| REML / GREML | In realistic samples the bias is usually upward, but in very large samples GREML can drift toward the random-mating baseline because the AM-induced long-range LD is not well captured by $\mathbf{G}$ |
| Relatives / family methods | Relatives become more genetically similar, so family-based estimators are affected too; the bias is method dependent |
| Cross-trait generalization | Score correlation can exceed effect correlation under xAM, inflating estimated $r_g$ |

::: {.callout-note}
Before Part 3, the big picture is: Part 1 and Part 2 mainly changed associations through **correlated genotypes**. Next we turn to **correlated environments**.
:::

# Part 3: Gene-Environment Correlations {background-color="#2c3e50" style="color: white;"}

## Same regression, new covariance sources

Now set the LD term aside and focus on gene-environment correlation:

$$
\tilde{\beta}_j = \beta_j + \frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}
$$

Part 3 is about a new question: **What can generate $\text{Cov}(x_j, e)$?**


::: {.callout-note}
In Part 1 and Part 2, the extra terms came mainly from **correlated genotypes**. Here the focus is **correlated environments**.
:::

## Where can $\text{Cov}(x_j,e)$ come from?
::: {style="font-size: 0.8em;"}
| Source | Why genotype and environment become correlated |
|--|----|
| Indirect genetic effects | Related people's genotypes shape the focal person's environment |
| Stratification / population structure | Allele frequencies and environments differ across groups |
| Residual confounding | Selection, ascertainment, or measurement structure |
| Assortative mating | Induced correlation between paternal genotype and maternal environment (and vice versa) |
:::
::: {.callout-note}
Different mechanisms can produce the **same statistical symptom** in a marginal GWAS coefficient: $\text{Cov}(x_j,e) \neq 0$.
:::
::: {.callout-important}
## Only passive gene-environment correlation is confounding
- **Passive $r_{ge}$**: Genotype and environment are correlated because of shared causes (e.g., parents)
- **Active / Evocative $r_{ge}$**: Genotype and environment are correlated because the focal person selects, creates, or evokes their environment based on their genotype (e.g., peers, partners, neighborhoods)

Crucially, **active and evocative gene-environment correlations are not confounding**. They are part of the causal pathway from genotype to phenotype, and are therefore mediators of the genetic effect. This is also important for how "direct" genetic effects are interpreted.

:::

## Indirect genetic effects

A focal person's phenotype can be affected not only by their own genotype, but also by the genotypes of parents, siblings, peers, or partners through the environments those people create.

$$
y_i = g_i^{\text{direct}} + g_i^{\text{indirect}} + e_i
\qquad
g_i^{\text{direct}} = \sum_{j=1}^{M}\beta_j x_{ij}
$$

:::: {.columns}
::: {.column width="50%"}
```{dot}
//| fig-width: 5
//| fig-height: 3.1
//| echo: false
digraph ige {
  rankdir=LR;
  graph [margin=0.08, nodesep=0.45, ranksep=0.7];
  node [shape=circle, style=filled, fixedsize=true, width=0.85, height=0.85,
        fontsize=17, fontcolor=white, fontname="Helvetica-Bold", penwidth=2];
  edge [penwidth=2, fontsize=15, fontname="Helvetica-Bold"];

  xr [label="x_rel", fillcolor="#e74c3c", color="#c0392b"];
  env [label="env", fillcolor="#e67e22", color="#ca6f1e"];
  xi [label="x_i", fillcolor="#3498db", color="#2980b9"];
  yi [label="y_i", fillcolor="#7f8c8d", color="#707b7c"];

  xr -> env [label=" indirect"];
  env -> yi [label=" environment"];
  xi -> yi [label=" direct"];
}
```
:::
::: {.column width="50%"}

- Your own genotype can affect your phenotype directly
- Other people's genotypes can affect your phenotype indirectly by shaping your environment
- That is why indirect genetic effects are not the same thing as stratification

:::
::::

## Stratification and population structure

Stratification can also induce correlation between genotype and environment, but through group-level differences rather than through other people's genotypes creating the focal environment.

```{dot}
//| fig-width: 5.2
//| fig-height: 3
//| echo: false
digraph strat {
  rankdir=TB;
  graph [margin=0.08, nodesep=0.5, ranksep=0.7];
  node [shape=circle, style=filled, fixedsize=true, width=0.9, height=0.9,
        fontsize=17, fontcolor=white, fontname="Helvetica-Bold", penwidth=2];
  edge [penwidth=2, fontsize=15, fontname="Helvetica-Bold"];

  anc [label="group", fillcolor="#8e44ad", color="#7d3c98"];
  xj  [label="x_j", fillcolor="#3498db", color="#2980b9"];
  e   [label="e", fillcolor="#e67e22", color="#ca6f1e"];
  y   [label="y", fillcolor="#7f8c8d", color="#707b7c"];

  anc -> xj;
  anc -> e;
  xj -> y [label=" marginal"];
  e  -> y [label=" env"];
}
```

::: {.callout-important}
Stratification is **confounding**, not nurture. It can create $\text{Cov}(x_j,e)$ even when there is no within-family environmental pathway from genotype to phenotype.
:::

## Back to the marginal SNP coefficient {.smaller}

For this part, the relevant simplification is:

$$
\tilde{\beta}_j
= \beta_j + \frac{\text{Cov}(x_j, e)}{\text{Var}(x_j)}
$$

::: {.callout-note}
The orange term can now reflect many different things:
- passive, evocative, or active gene-environment correlation
- indirect genetic effects from parents, siblings, peers, or partners
- stratification and population structure
- other residual confounding
:::

## From SNP confounding to variance decomposition

For the next two slides, let $g$ denote the focal individual's **direct additive genetic value**:

$$
g_i = \sum_{j=1}^{M} \beta_j x_{ij}
$$

Then:

$$
y = g + e
\qquad \Longrightarrow \qquad
\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)
$$

::: {.callout-note}
If you split indirect genetic effects out as their own term, you get more covariance terms, not fewer. The interpretational problem does not disappear.
:::

## Why covariance terms complicate interpretation

Keep these quantities conceptually separate:

$$
\text{Var}(g) \qquad \text{Var}(e) \qquad \text{Cov}(g,e) \qquad \text{Var}(y)
$$

The sign of the covariance matters:

| Sign of $\text{Cov}(g,e)$ | Interpretation |
|---|---|
| Positive | Genetically higher-valued individuals tend to experience environments that push in the same direction |
| Negative | Genetically higher-valued individuals tend to experience environments that push in the opposite direction |

::: {.callout-important}
$\text{Var}(y) = \text{Var}(g) + \text{Var}(e) + 2\text{Cov}(g,e)$ is a **statistical identity**, not a clean causal partition of phenotypic variation.
:::

## Take-home map

| Quantity | What it means | What it does not automatically mean |
|---|---|---|
| Direct effect $\beta_j$ | Causal effect of locus $j$ in the full model | The same thing as the marginal coefficient |
| Marginal coefficient $\tilde{\beta}_j$ | Association between SNP $j$ and phenotype | A pure direct effect |
| Indirect genetic effect | Other people's genotypes shape the focal environment | Population stratification |
| Stratification | Group structure aligns genotypes and environments | Genetic nurture |
| $\text{Cov}(g,e)$ | Statistical alignment of genetic values and environments | A causal share of variance |

::: {.callout-note}
Across local LD, assortative mating, and gene-environment correlation, the same lesson keeps returning: **marginal regression coefficients are shaped by correlation structure**.
:::

# Part 4: Family-based designs (overview) {background-color="#2c3e50" style="color: white;"}

## Why bring up families at all?

Much of this lecture has been about one central problem: the **population** coefficient $\tilde{\beta}_j$ is not always the same as the direct effect $\beta_j$.

Across unrelated people, SNP associations can reflect:

- local LD
- assortative mating
- stratification
- parental genetic nurture and other shared family environments

::: {.callout-note}
Family-based designs matter because they change the comparison. Instead of asking why people from different families differ, they compare relatives who share much of their background.
:::

## Core intuition

Within a family, many background features are held much more constant:

- the parents are the same
- ancestry background is much more similar
- much of the rearing environment is shared

What still differs is the random segregation of alleles from parents to offspring.

Recall from Session 1:

$$
x_{jc} =
\underbrace{\left(\frac{1}{2}x_{jp} + \epsilon_{jp}\right)}_{\pcol{x_j^{p}}}
+
\underbrace{\left(\frac{1}{2}x_{jm} + \epsilon_{jm}\right)}_{\mcol{x_j^{m}}}
$$

$$
\text{Var}(x_{jc})
= \frac{1}{4}\text{Var}(x_{jp}) + \frac{1}{4}\text{Var}(x_{jm}) + \frac{1}{2}\text{Cov}(x_{jp}, x_{jm})
+ \underbrace{\text{Var}(\epsilon_{jp} + \epsilon_{jm})}_{\text{segregation variance}}
$$

::: {.callout-note}
This is the callback to Session 1: the **random segregation deviations** do the heavy lifting. The key term is the segregation variance, and it is **random**. It comes from which alleles happened to be transmitted, not from stable differences in ancestry, parents, or shared family background.
:::

::: {.callout-note}
Across siblings, the parents are the same, so much of the family background is fixed. What differs is the realized draw of the $\epsilon$ terms. That is why family designs use **within-family random segregation** to get closer to direct effects than ordinary population GWAS.
:::

## What family designs are good for

| Question | High-level answer |
|---|---|
| Why can they help with stratification? | Relatives are compared against each other rather than against people from different ancestry backgrounds |
| Why can they help with parental genetic nurture? | Siblings share the same parents, so much of the parental background is held fixed |
| Why can they help with assortative mating? | They hold the parental pair fixed, removing much of the between-family assortment signal |
| Why are they appealing conceptually? | They target the same confounding problem we have discussed all day, but by changing the design rather than adding more covariates |

## What they do **not** magically solve

Family-based estimates are still not automatically equal to the direct causal effect.

| Remaining limitation | Why it remains |
|---|---|
| Local and long-range LD | Relatives still inherit correlated variants, including background correlation induced by assortative mating |
| Indirect effects within families | One relative's genotype can still affect another relative's environment |
| Non-shared environments | Siblings do not experience identical environments |
| Lower precision | Within-family contrasts use less genetic variation than population GWAS |

::: {.callout-important}
The right takeaway is not "family GWAS is perfect." The right takeaway is that it removes much of the **between-family confounding** that worries us in population GWAS.
:::

## Cleaner weights are not the same as unconfounded scores

Family designs mainly help with the **estimated SNP weight** by making the marginal coefficient closer to a direct effect.

But many downstream uses depend on both the weight and the realized genotype:

$$
\mathrm{PGS}_i^{F} = \sum_j \hat{\tilde{\beta}}_j^{F} x_{ij}
$$

That does **not** mean the score itself is automatically unconfounded.

Two problems can still remain at the SNP level:

- $x_{ij}$ can still be correlated with other causal SNPs through local or long-range LD
- $x_{ij}$ can still be correlated with environments, so $\text{Cov}(x_{ij}, e_i)$ need not vanish

::: {.callout-important}
Even a perfectly measured genetic value $g$ would still be associated with the phenotype through both direct and indirect pathways. A polygenic score has the same problem, so better SNP weights alone do not make it unconfounded.
:::

## Why adjust for parental or sibling polygenic scores?

At the polygenic score level, the same segregation logic still applies. A focal person's score can be written as:

$$
\mathrm{PGS}_F = \frac{1}{2}\left(\mathrm{PGS}_p + \mathrm{PGS}_m\right) + \Delta_F
$$

where:

- $\frac{1}{2}(\mathrm{PGS}_p + \mathrm{PGS}_m)$ is the expected family background in the score
- $\Delta_F$ is the **random segregation deviation** in the score

If the focal phenotype depends both on the focal genotype and on family-level genetic background, then regressing $y_F$ on $\mathrm{PGS}_F$ alone will mix those two sources of variation.

::: {.callout-note}
So the problem is the same as before: the observed polygenic score contains both a **between-family component** and a **within-family segregation component**.
:::

## Why the adjustment helps

If we fit

$$
y_F = b_1 \mathrm{PGS}_F + b_2 \overline{\mathrm{PGS}}_{parent} + \varepsilon_F
\qquad \text{or} \qquad
y_F = b_1 \mathrm{PGS}_F + b_2 \mathrm{PGS}_{sib} + \varepsilon_F
$$

then the extra family score absorbs part of the shared family-genetic background.

That means the remaining identifying variation in $\mathrm{PGS}_F$ is closer to the segregation term $\Delta_F$:

- adjusting for **parental** scores targets the family background more directly
- adjusting for a **sibling** score uses another child of the same parents as a proxy for that shared background
- in both cases, the coefficient on $\mathrm{PGS}_F$ becomes more within-family in spirit

::: {.callout-important}
This is why these controls can help: they soak up the part of the focal score that is predictable from the family, leaving more of the signal that comes from the focal person's **random segregation draw**.
:::

::: {.callout-note}
But again, this is not magic. LD, residual $\text{Cov}(x,e)$, sibling indirect effects, and measurement error in the scores can still matter.
:::

## Mixed-model family designs

Sibling comparisons are one kind of family design. Another is to use relatives inside a **mixed model** to separate direct and indirect genetic pathways.

## Intuition: ordinary GREML vs M-GCTA

Ordinary GREML asks:

- Are focal people who are slightly more genetically similar also slightly more phenotypically similar?

M-GCTA asks:

- Are focal people whose **mothers** are slightly more genetically similar also slightly more phenotypically similar?

::: {.callout-note}
In both cases the logic is covariance matching. If a source of genetic similarity creates phenotype similarity, give the mixed model the corresponding covariance matrix.
:::

## The M-GCTA model

In Session 1, GREML used a GRM to model covariance from individuals' own genotypes:

$$
\mathbf{y}_{F} = \mathbf{g}_{F} + \mathbf{e}_{F},
\qquad
\mathbf{g}_{F} \sim \mathcal{N}(\mathbf{0},\; \sigma_F^2 \mathbf{G}_{F})
$$

Here $F$ just means the **focal phenotype**. It could be an offspring phenotype, a maternal phenotype, or any other phenotype measured in one family member.

If a relative's genotype also helps shape that focal phenotype, then part of what we are calling "environment" has its own genetic covariance structure:

$$
\mathbf{y}_{F} = \mathbf{g}_{F} + \mathbf{g}_{R \rightarrow F} + \mathbf{e}_{F}
$$

::: {.callout-note}
The logic is exactly the same as ordinary GREML: if a pathway creates phenotype covariance, give the model the covariance matrix for that pathway. The only difference is that we now include covariance from a **relative's** genotype as well as the focal person's own genotype.
:::

In today's practical, the focal phenotype is the offspring and the relevant relative is the mother. So M-GCTA extends the direct-only GREML model by adding:

- a focal direct component based on $\mathbf{G}_{F} = \mathbf{G}_{oo}$
- a maternal indirect component based on $\mathbf{G}_{R} = \mathbf{G}_{mm}$
- a covariance component for overlap between direct and maternal effects

::: {.callout-important}
Following Eilertsen et al. (2021), this logic is not inherently tied to an offspring phenotype. The framework is defined relative to a **focal phenotype** and can be applied to phenotypes measured in different family members.
:::

::: {.callout-important}
This is why M-GCTA belongs in the family-design toolbox: it uses known family roles and multiple related genomes to ask whether a relative's genotype shapes the focal outcome through the environment, rather than forcing all genotype-related covariance into the focal person's direct effect.
:::

## Bridge to tomorrow

For this session, the main message is just:

$$
\text{Population GWAS: across-family variation}
\qquad \longrightarrow \qquad
\text{Family designs: within-family variation}
$$

That shift in identifying variation is why family-based methods are such an important complement to standard GWAS.

::: {.callout-note}
We will leave the design details, estimators, and tradeoffs for the family-based session, and keep today's treatment at the conceptual level.
:::

# Part 5: Practical {background-color="#2c3e50" style="color: white;"}

## Practical 1: GREML and M-GCTA with simulated trios

Session 2 continues the same `practical/sim_genomic.R` (or `practical/sim_genomic.jl`) script used in Session 1.

Today's new extension is:

1. Start from the direct-only setup already covered in Session 1
2. Simulate an offspring phenotype with a maternal indirect genetic effect
3. Refit the wrong direct-only GREML model
4. Fit the maternal model (M-GCTA) and compare the variance components

::: {.callout-note}
Parts 1-4 of the script are the same setup from Session 1, so they are not repeated on the slides here. If needed, rerun them in the script to recreate `X_tilde`, `yo`, `G`, `Goo`, and the direct-only GREML fit.
:::

## Starting point from Session 1 {.smaller}

By this point, you should already have the direct-only objects from the Session 1 practical:

- standardized genotypes `X_tilde`
- offspring phenotype `yo`
- genomic relationship matrix `G`
- offspring block `Goo`
- direct-only GREML fit

The slides below focus only on the **new Session 2 extension**.

```{r}
#| label: s2-session1-setup
#| echo: false
#| include: false
#| cache: true
set.seed(080318)
library(gaston)

N = 3000
M = 12000
p = 0.5
Xf = matrix(rbinom(N * M, 2, p), N, M)
Xm = matrix(rbinom(N * M, 2, p), N, M)
Xo = matrix(
  rbinom(N * M, 1, as.vector(Xf) / 2) +
  rbinom(N * M, 1, as.vector(Xm) / 2), N, M)
X = rbind(Xf, Xm, Xo)

idf = 1:N
idm = (N + 1):(2 * N)
ido = (2 * N + 1):(3 * N)

p_est = colMeans(X) / 2
X_tilde = scale(X, 2 * p_est, sqrt(2 * p_est * (1 - p_est)))
s2_g = 0.5
s2_e = 0.5
b = rnorm(M, 0, sqrt(s2_g / M))
g = as.vector(X_tilde %*% b)
e = rnorm(3 * N, 0, sqrt(s2_e))
y = g + e

yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
m_po = lm(yo ~ yp)

G = tcrossprod(X_tilde) / M
Goo = G[ido, ido]
m_direct = lmm.aireml(yo, K = Goo, verbose = FALSE)
```

```julia
#| echo: false
#| eval: false
using LinearAlgebra, Distributions, DataFrames, GLM
using GREMLModels, Random
Random.seed!(080318)

N, M = 3000, 12000
p = 0.5
Xf = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xm = rand(Bernoulli(p), N, M) .+ rand(Bernoulli(p), N, M)
Xo = rand.(Bernoulli.(Xf ./ 2)) .+ rand.(Bernoulli.(Xm ./ 2))
X = vcat(Xf, Xm, Xo)

idf = 1:N
idm = N+1:2N
ido = 2N+1:3N

p_est = vec(mean(X, dims = 1)) ./ 2
X_tilde = (Float64.(X) .- (2 .* p_est)') ./ sqrt.(2 .* p_est .* (1 .- p_est))'

sigma2_g, sigma2_e = 0.5, 0.5
b = randn(M) * sqrt(sigma2_g / M)
g = X_tilde * b
e = randn(3N) * sqrt(sigma2_e)
y = g + e

yp = 0.5 * (y[idf] + y[idm])
yo = y[ido]
df1 = DataFrame(yo = yo, yp = yp)
mpo = lm(@formula(yo ~ yp), df1)

G = (X_tilde * X_tilde') / M
Goo = G[ido, ido]
Id = Diagonal(ones(N))
r = [Goo, Id]
mdirect = fit(GREMLModel, @formula(yo ~ 1), df1, r; verbose = false)
```

## Part 5: Simulate a maternal indirect genetic effect {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-maternal-sim
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-session1-setup"
Sg = rbind(
  c(0.2, 0.2),
  c(0.2, 0.5)
)
s2_e = 0.1
B = matrix(rnorm(2 * M), M, 2) %*% chol(Sg / M)
g_m = X_tilde %*% B[, 1]
g_d = X_tilde %*% B[, 2]
e_2 = g_m[idm] + rnorm(N, 0, sqrt(s2_e))
yo_2 = g_d[ido] + e_2
```

### Julia
```julia
Sigma_g = [0.2 0.2; 0.2 0.5]
sigma2_e = 0.1
C = cholesky(Symmetric(Sigma_g / M)).U
B = randn(M, 2) * C
g_m = X_tilde * B[:, 1]
g_d = X_tilde * B[:, 2]
e2 = g_m[idm] + randn(N) * sqrt(sigma2_e)
yo2 = g_d[ido] + e2
```

::::

## Part 6: Fit the wrong direct-only model {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-wrong-direct
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-maternal-sim"
m_direct_2 = lmm.aireml(yo_2, K = Goo, verbose = FALSE)
m_direct_2$tau
```

### Julia
```julia
df2 = DataFrame(y = yo2)
mdirect2 = fit(GREMLModel, @formula(y ~ 1), df2, r; verbose = false)
```

::::

## Part 7: Fit the maternal model (M-GCTA) {.smaller}

:::: {.panel-tabset}

### R
```{r}
#| label: s2-mgcta
#| echo: true
#| code-fold: false
#| output-location: column
#| cache: true
#| dependson: "s2-wrong-direct"
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
m_maternal = lmm.aireml(yo_2, K = list(Gmm, Dom, Goo), verbose = FALSE)
m_maternal$tau
```

### Julia
```julia
Gmm = G[idm, idm]
Dom = G[ido, idm] + G[idm, ido]
r2 = [Gmm, Dom, Goo, Id]
mmaternal = fit(GREMLModel, @formula(y ~ 1), df2, r2; verbose = false)
```

::::

## Interpretation checkpoint

After running the simulation, check that you can answer these questions:

1. Does parent-offspring regression recover something close to the simulated `h^2 = 0.5` in the direct-only setting?
2. In the direct-only GREML model, what do the estimated variance components look like before and after adding the maternal indirect effect?
3. Once you fit the maternal model, do the variance components move closer to the data-generating values?
4. Why does the direct-only model misattribute part of the maternal effect?

## Practical 2: LD score regression with real GWAS data

In this practical, we will use **LD score regression (LDSC)** to estimate:

1. observed-scale SNP heritability for **intelligence**
2. observed-scale SNP heritability for **p_factor**
3. the genetic correlation between the two traits

The summary statistics come from the **CTG summary statistics archive** and the practical uses a local course installer around the newer Python 3 LDSC workflow from **CBIIT/ldsc**. LDSC itself remains the canonical method originally associated with the **bulik/ldsc** project.

- [CTG GWAS summary statistics](https://ctg.thebluebus.nl/software/summary_statistics/)
- [CBIIT/ldsc repository](https://github.com/CBIIT/ldsc)
- [Original bulik/ldsc repository](https://github.com/bulik/ldsc)
- [LDSC heritability and genetic correlation tutorial](https://github.com/bulik/ldsc/wiki/Heritability-and-Genetic-Correlation)

::: {.callout-note}
To keep the exercise focused, the instructor has already prepared helper scripts for the repetitive setup work. You will still run the same main LDSC steps used in practice: `munge_sumstats.py`, single-trait `--h2`, and cross-trait `--rg`, but the course scripts handle the local installation and platform details.
:::

## Setup and inputs {.smaller}

We will work with two pre-harmonized files prepared from the public CTG releases:

- `practical/ldsc/data/course/intelligence_ctg.tsv.gz`
- `practical/ldsc/data/course/p_factor.tsv.gz`

Each file has the same minimal schema:

`SNP`, `A1`, `A2`, `P`, `N`, `Z`

We also need the LDSC reference files:

- `practical/ldsc/data/ref/w_hm3.snplist`
- `practical/ldsc/data/ref/eur_w_ld_chr/`
- `practical/ldsc/data/ref/weights_hm3_no_hla/`

Run the course workflow in this order from the repository root. The practical is intended to work in **Git Bash on Windows**, with the scripts handling the local Python and LDSC setup.

```bash
Rscript practical/ldsc/prepare_ldsc_data.R --force
bash practical/ldsc/install_ldsc.sh
bash practical/ldsc/ldsc_pipeline.sh
```

::: {.callout-note}
These LD scores are based on **1000 Genomes Europeans**, so the exercise assumes European-ancestry GWAS summary statistics. In real work, ancestry mismatch between the GWAS and the reference panel can bias interpretation.
:::

## Part 1: Munge the summary statistics {.smaller}

After running the pipeline above, the first LDSC step is to convert the prepared text files into LDSC's `.sumstats.gz` format. If you want to inspect the script itself, see `practical/ldsc/ldsc_pipeline.sh`.

Look for:

- how many SNPs survive QC and allele matching for `intelligence` and `p_factor`
- whether the munge logs contain warnings
- the two output files `intelligence_ctg.sumstats.gz` and `p_factor.sumstats.gz`

## Part 2: Estimate SNP heritability {.smaller}

That same pipeline run then performs single-trait LDSC for each phenotype.

Look for these lines in each log:

- `Total Observed scale h2`
- `Intercept`
- `Ratio`

The key log files are:

- `practical/ldsc/data/results/intelligence_h2.log`
- `practical/ldsc/data/results/p_factor_h2.log`

::: {.callout-note}
For this short practical we stay on the **observed scale**. We are not doing liability-scale conversion or partitioned heritability here.
:::

## Part 3: Estimate the genetic correlation {.smaller}

Finally, that same pipeline run performs the cross-trait LDSC regression using the two munged `.sumstats.gz` files.

Look for:

- `Genetic Correlation`
- its standard error
- the cross-trait intercept
- the `P:` line

The main log file is:

- `practical/ldsc/data/results/intelligence_p_factor_rg.log`

For these traits, we expect a **negative** estimated genetic correlation.

::: {.callout-note}
After Part 2, keep one interpretation caveat in mind: an LDSC estimate of $r_g$ from marginal summary statistics is not automatically pure effect overlap / pleiotropy if cross-trait assortative mating or other covariance structure is present.
:::

## Interpretation checkpoint

After the run, check that you can answer these questions:

1. Which trait has the larger observed-scale SNP heritability: `intelligence` or `p_factor`?
2. Are the LDSC intercepts close to 1, or do they suggest substantial inflation beyond polygenicity?
3. Are the LDSC ratios modest, or do they suggest a large share of inflation is due to confounding?
4. Is the estimated genetic correlation between `intelligence` and `p_factor` negative, and how large is it in magnitude?
5. Is the cross-trait intercept close to 0, or could sample overlap / shared confounding be contributing?

::: {.callout-important}
The important habit is not just to report `h^2` and `r_g`, but also to inspect the **intercepts** and **ratios** and ask whether the analysis assumptions look reasonable.
:::





